### 2022-09-20 notes/notes.tex
first version of notes

\documentclass{article}
\usepackage{amsmath, amssymb, dsfont, physics}
\usepackage[margin=1.5cm]{geometry}

\title{Notes}
\author{SCK team}
\date{}

\begin{document}
    \maketitle

\section{Notation}
\begin{itemize}
    \item Average emission per source: $x_1,\dots,x_m$ where $m = 200$.
    \item Measured concentrations $y_1,\dots,y_n$ where $n = 4636$. These are consecutive measurements for several measurement stations. 
    \item $M_{i,(j,\Delta)}$ is a block in the sensitivity matrix $M$ that represents the contribution of $x_j$ from $\Delta - 1$ days ago to the observed concentration $y_i$.
    \item Scaling factors $s_1,\dots,s_m$ (in \texttt{scalings.csv}). These are put in a diagonal matrix $S$.
    \item $\vb{1}_q = [1\,\, \dots\,\, 1]^T \in \mathbb{R}^q$ is the constant vector consisting of only ones.
    \item $\otimes$ is the Kronecker product. In particular,
        $$
        \underbrace{
        \begin{bmatrix}
            a_1 \\
            \vdots \\
            a_p
        \end{bmatrix}
        }_{\in \mathbb{R}^{p}}
        \otimes 
        \underbrace{
        \begin{bmatrix}
            1 \\
            \vdots \\
            1
        \end{bmatrix}
        }_{\in \mathbb{R}^{q}}
        =
        \underbrace{
        \begin{bmatrix}
            a_1 \\
            \vdots \\
            a_1 \\
            a_2 \\
            \vdots \\
            a_2 \\
            \vdots \\
            a_p
        \end{bmatrix}
        }_{\in \mathbb{R}^{pq}}
        .$$
\end{itemize}

\section{Model}
With $n=1,\dots,4636$ and $m = 1,\dots,200$, the initial model is
$$
y_i \approx
\hat{y}_i = \sum_{j=1}^m \sum_{\Delta = 1}^{15}  M_{i,(j,\Delta)} \frac{x_{j}}{s_j}
.$$
That is, for each emission source $j$, we simulate the contribution of 15 days, and add this together for all $j$. A matrix formulation of this reads

\begin{align*}
\begin{bmatrix}
    \hat{y}_1 \\ 
    \vdots \\
    \hat{y}_{m}
\end{bmatrix}
&=
\begin{bmatrix}
    M_{1,(1,1)} & M_{1,(1,2)} & \cdots & M_{1,(n,15)} \\
    \vdots & \vdots & & \vdots \\
    M_{m,(1,1)} & M_{m,(1,2)} & \cdots & M_{m,(n,15)} \\
\end{bmatrix}
\begin{bmatrix}
    s_1^{-1} x_1\\
    \vdots \\
    s_1^{-1} x_1 \\
    s_2^{-1} x_2 \\
    \vdots \\
    s_n^{-1} x_n
\end{bmatrix} \\
&=
\begin{bmatrix}
    M_{1,(1,1)} & M_{1,(1,2)} & \cdots & M_{1,(n,15)} \\
    \vdots & \vdots & & \vdots \\
    M_{m,(1,1)} & M_{m,(1,2)} & \cdots & M_{m,(n,15)} \\
\end{bmatrix}
\left( 
     \begin{bmatrix}
        s_1 & & \\
         & \ddots & \\
        & & s_n
     \end{bmatrix}^{-1}
     \begin{bmatrix}
        x_1 \\
        \vdots \\
        x_n
     \end{bmatrix}
     \otimes \vb{1}_{15}
\right)
\end{align*}
where $s_i$ are the scalings.

Now we present a different interpretation. Instead of $S^{-1} x \otimes \vb{1}_{15}$, we make a vector $\tilde{x} := S^{-1} x \otimes \vb{1}_{365}$. This is a vector of length $365 \times 200$ that gives the emission from day 1 to day 365 for each source. This is assumed to be constant for all days. Therefore, $\tilde{x}$ consists of $m = 200$ blocks, each of which is constant.

With this formulation, the predicted concentration in station $y_i$ is 
$$
        \hat{y}_i
    =
    \begin{bmatrix}
        \cdots & M_{1,(1,1)} & \dots & M_{1,(1,15)} & 0 & \dots & M_{1,(2,1)} & \dots & M_{1,(2,15)} & \dots
    \end{bmatrix}
    (S^{-1} x \otimes \vb{1}_{365})
.
$$

Now assume that in each station, there is one observation per day, named $y_{kt}$, where $k = 1,\dots,K$ where $K$ is the number of stations and $t = 1,\dots,365$. The index $t=1$ corresponds to the most recent measurement. Then 
$$
\begin{bmatrix}
    \hat{y}_{1,1} \\
    \hat{y}_{1,2} \\
    \vdots \\
    \hat{y}_{2,1} \\ 
    \vdots \\
    \hat{y}_{K,365}
\end{bmatrix}
=
\underbrace{
\begin{bmatrix}
    M_{(1,1), (1,1)} & M_{(1,1), (1,2)} & \cdots & M_{(1,1),(1,15)} & 0 & \cdots & \cdots & M_{(1,1), (2,1)} & \cdots & 0 \\
    0 & M_{(1,2), (1,1)} & \cdots & M_{(1,2), (1,14)} & M_{(1,2), (1,15)} & 0 & \cdots & 0 & \cdots & 0 \\
    \vdots & \vdots & & \vdots & \vdots & \vdots & & \vdots & & \vdots \\
    M_{(2,1), (1,1)} & M_{(2,1), (1,2)} & \cdots & M_{(2,1),(1,15)} & 0 & \cdots & \cdots & M_{(2,1), (2,1)} & \cdots & 0 \\
    \vdots \\
    \vdots
\end{bmatrix}
}_{ := M_{shift}}
\tilde{x}
$$
where 
$$
\tilde{x} := S^{-1} x \otimes \vb{1}_{365}
=
\begin{bmatrix}
    s_1^{-1} x_1 \\
    \vdots \\
    s_1^{-1} x_1 \\ 
    s_2^{-1} x_2 \\
    \vdots \\ 
    s_m^{-1} x_m
\end{bmatrix}
.$$
If we weigh each day and each emission source by some factor, we get
$$
\hat{y} = M_{shift} W (S^{-1} x \otimes \vb{1}_{365})
$$
where $W$ is a diagonal matrix containing the weights.
If $W$ is what we should be looking for, then we can use well-known numerical techniques for solving linear matrix equations. For theoretical purposes, we describe the problem as a large linear system involving Kronecker products.
Writing $W = \mathrm{diag}(w)$ for some vector $w$, the above can be reformulated as a liner operation applied to $w$ as follows:
$$
\hat{y} = ((S^{-1} x \otimes \vb{1}_{365})^T \otimes M_{shift}) \mathrm{vec}(\mathrm{diag}\, w)
$$
where $\mathrm{vec} \circ \mathrm{diag}$ is a constant linear map that can be inferred from the dimensions of the problem.

\end{document}

### 2022-09-20 notes/notes.tex
clarification of theory

@@
 \documentclass{article}
-\usepackage{amsmath, amssymb, dsfont, physics}
+\usepackage{amsmath, amssymb, dsfont, physics, hyperref, cleveref}
 \usepackage[margin=1.5cm]{geometry}
+\usepackage{biblatex}
 
@@
 \date{}
+
+\addbibresource{library.bib}
 
@@
         .$$
+        A standard rule of thumb in numerical linear algebra is that the Kronecker product serves a mostly theoretical purpose and is rarely calculated explicitly.
+See, e.g.
+\cite[Section 12.3]{Golub2013}.
 \end{itemize}
@@
 \section{Model}
+\subsection{Initial model}
 With $n=1,\dots,4636$ and $m = 1,\dots,200$, the initial model is
@@
 .$$
-That is, for each emission source $j$, we simulate the contribution of 15 days, and add this together for all $j$. A matrix formulation of this reads
-
-\begin{align*}
+That is, for each emission source $j$, we simulate the contribution of 15 days, and add this together for all $j$. A matrix formulation of this reads as follows:
+\begin{align}
+    \label{eq: initial model concise} 
 \begin{bmatrix}
@@
     \vdots \\
-    \hat{y}_{m}
+    \hat{y}_{n}
 \end{bmatrix}
@@
 \begin{bmatrix}
-    M_{1,(1,1)} & M_{1,(1,2)} & \cdots & M_{1,(n,15)} \\
+    M_{1,(1,1)} & M_{1,(1,2)} & \cdots & M_{1,(m,15)} \\
     \vdots & \vdots & & \vdots \\
-    M_{m,(1,1)} & M_{m,(1,2)} & \cdots & M_{m,(n,15)} \\
+    M_{n,(1,1)} & M_{n,(1,2)} & \cdots & M_{n,(m,15)} \\
 \end{bmatrix}
@@
     \vdots \\
-    s_n^{-1} x_n
-\end{bmatrix} \\
+    s_n^{-1} x_m
+\end{bmatrix} 
+\\
 &=
@@
 \right)
-\end{align*}
-where $s_i$ are the scalings.
+\nonumber
+.\end{align}
 
 Now we present a different interpretation.
-Instead of $S^{-1} x \otimes \vb{1}_{15}$, we make a vector $\tilde{x} := S^{-1} x \otimes \vb{1}_{365}$.
+Instead of $S^{-1} x \otimes \vb{1}_{15}$, we make a vector $\tilde{x}_{365} := S^{-1} x \otimes \vb{1}_{365}$.
 This is a vector of length $365 \times 200$ that gives the emission from day 1 to day 365 for each source.
 This is assumed to be constant for all days.
-Therefore, $\tilde{x}$ consists of $m = 200$ blocks, each of which is constant.
+Therefore, $\tilde{x}_{365}$ consists of $m = 200$ blocks, each of which is a constant vector of length $365$.
 
-With this formulation, the predicted concentration in station $y_i$ is 
+We can write a system of equations that is logically the same as \cref{eq: initial model concise} but uses $\tilde{x}_{365}$.
+This gives a larger linear system.
+Because the emission from most dates has a negligible contribution to the observed concentrations, many coefficients in the larger matrix will be zero.
+For every measurement $i$, we have a linear relationship that looks something like this:
 $$
@@
     \begin{bmatrix}
-        \cdots & M_{1,(1,1)} & \dots & M_{1,(1,15)} & 0 & \dots & M_{1,(2,1)} & \dots & M_{1,(2,15)} & \dots
+        \cdots & M_{i,(1,1)} & \dots & M_{i,(1,15)} & 0 & \dots & M_{i,(2,1)} & \dots & M_{i,(2,15)} & \dots
     \end{bmatrix}
@@
 $$
+The presence of zeros in the matrix indicates that emissions from the distant past do not contribute to the observed concentrations.
 
@@
 The index $t=1$ corresponds to the most recent measurement.
-Then
-$$
+Then the full system reads as follows:
+\begin{equation}
+    \label{eq: initial model 365 days}
 \begin{bmatrix}
@@
 }_{ := M_{shift}}
-\tilde{x}
-$$
+\tilde{x}_{365}
+\end{equation}
 where 
 $$
-\tilde{x} := S^{-1} x \otimes \vb{1}_{365}
+\tilde{x}_{365} := S^{-1} x \otimes \vb{1}_{365}
 =
@@
 .$$
-If we weigh each day and each emission source by some factor, we get
+
+We could reorganise the vector $\tilde{x}_{365}$ to the transposed expression $\vb{1}_{365} \otimes (S^{-1} x)$, which is a vector containing all emissions on day 1, followed by all emissions on day 2, etc.
+A similar permutation of $y$ could be chosen.
+This has the effect of permuting the rows and columns of $M_{shift}$.
+A permutation of the variables gives a model that is essentially the same as \cref{eq: initial model 365 days} but may have advantages when it comes to implementation or computational efficiency.
+We will not consider this.
+
+\subsection{Model with time-dependent correction to the emissions}
+If we weigh each day and each emission source by some factor, we are applying a transformation
 $$
+\begin{bmatrix}
+    s_1^{-1} x_1 \\
+    \vdots \\
+    s_1^{-1} x_1 \\ 
+    s_2^{-1} x_2 \\
+    \vdots \\ 
+    s_m^{-1} x_m
+\end{bmatrix}
+\mapsto
+\begin{bmatrix}
+    w_{1,1} s_1^{-1} x_1 \\
+    \vdots \\
+    w_{1,365} s_1^{-1} x_1 \\ 
+    w_{2,1} s_2^{-1} x_2 \\
+    \vdots \\ 
+    w_{m, 365} s_m^{-1} x_m
+\end{bmatrix}
+$$
+which can be written concisely as 
+$$
+W (S^{-1} x \otimes \vb{1}_{365})
+\quad\text{where}\quad 
+W = \begin{bmatrix}
+    w_{1,1} & 0 & \cdots & 0 \\
+    0 & w_{1,2} & \cdots & 0 \\
+    \vdots & \vdots & \ddots & 0 \\
+    0 & 0 & \cdots & w_{m, 365}
+\end{bmatrix}
+.$$
+Using this weighted vector of emissions instead of $\tilde{x}_{365}$ in \cref{eq: initial model 365 days} gives
+\begin{equation}
+    \label{eq: time-evolving model concise}
 \hat{y} = M_{shift} W (S^{-1} x \otimes \vb{1}_{365})
+.\end{equation}
+In this expression, the unknown is $W$ and should be fit to the data. 
+There are a plethora of numerical techniques for solving linear equations where the unknown is a matrix.
+In the following, we provide a theoretical description of a system that can be solved for $W$.
+
+We can put constraints on $W$ of the form $W \in \mathcal{W}$, where $\mathcal{W}$ is some set of constrained diagonal matrices.
+Suppose we have a parametrisation $\Gamma: \mathbb{R}^{k} \rightarrow \mathcal{W}, u \mapsto \Gamma(u)$.
+In this case, \cref{eq: time-evolving model concise} can be reformulated as a linear operation applied to $\Gamma(u)$ as follows:
 $$
-where $W$ is a diagonal matrix containing the weights.
-If $W$ is what we should be looking for, then we can use well-known numerical techniques for solving linear matrix equations.
-For theoretical purposes, we describe the problem as a large linear system involving Kronecker products.
-Writing $W = \mathrm{diag}(w)$ for some vector $w$, the above can be reformulated as a liner operation applied to $w$ as follows:
-$$
-\hat{y} = ((S^{-1} x \otimes \vb{1}_{365})^T \otimes M_{shift}) \mathrm{vec}(\mathrm{diag}\, w)
-$$
-where $\mathrm{vec} \circ \mathrm{diag}$ is a constant linear map that can be inferred from the dimensions of the problem.
+\hat{y} = ((S^{-1} x \otimes \vb{1}_{365})^T \otimes M_{shift}) \mathrm{vec}\, \Gamma(u)
+.$$
+If $\Gamma$ is a linear map, the above is a linear expression in $u$. Then, the minimisation of $\norm{y - \hat{y}}^2$ can be solved using numerical linear algebra.
+
+\printbibliography
 

### 2022-09-20 code/import.py
Dictionary for measurement files

@@ -43,6 +43,7 @@ print(nb_data_cols)
 
 combined_dataframe=pandas.DataFrame()
 
+measurements_dict = {};
 for file in os.listdir(MATRIX_LOCATION):
     if file.endswith(".csv"):
         matrix_file = pandas.read_csv(os.path.join(MATRIX_LOCATION, file), header=None)
@@ -56,7 +57,7 @@ for file in os.listdir(MATRIX_LOCATION):
         time = split_name[2]
 
         # print("matrix: ", np_matrix.shape)
-
+        measurements_dict.update({pruned_name:{"location": location, "date": date, "time": time, "matrix": np_matrix}})
 
 
         #checking whether everything sums to 1 (or something negative)
@@ -80,3 +81,4 @@ plt.xlabel("log10(y_sim)")
 plt.legend()
 plt.title("Simulated observations")
 plt.show()
+

### 2022-09-20 code/import.py
Constructed combined matrix

@@ -2,8 +2,16 @@ import pandas
 import os
 import numpy as np
 import matplotlib.pyplot as plt
+import scipy.sparse as sp
+import datetime
 
 MATRIX_LOCATION = "./../data/srs_csv_files/"
+#TODO: figure out these parameters from the files themselves
+LIMIT_BACKWARD_TIME = 15
+N_SOURCES = 200
+N_DETECTIONS = 4636
+#allowing us to use values for 1 entire year
+N_COLS=365+LIMIT_BACKWARD_TIME
 
 
 scalings=pandas.read_csv("./../data/scalings.csv", header=None)
@@ -24,9 +32,9 @@ np_sources=np_sources/np_scalings
 #np.divide(np_sources,np_scalings)
 
 
-print(np_scalings.size)
-print(np_scalings.shape)
-print("np sources: ", np_sources.size)
+# print(np_scalings.size)
+# print(np_scalings.shape)
+# print("np sources: ", np_sources.size)
 
 n_datafiles=0
 for file in os.listdir(MATRIX_LOCATION):
@@ -35,6 +43,19 @@ for file in os.listdir(MATRIX_LOCATION):
 results=np.zeros(n_datafiles)
 print(results)
 
+
+
+
+#TODO figure out whether this thing needs to be sparse
+#probably
+# def fill_matrix(y, index, ):
+    # row =
+    # col =
+    # data =
+#or just construct from standard numpy array...
+
+
+
 index=0
 
 #timesteps times number of emissions
@@ -42,6 +63,10 @@ nb_data_cols=np.size(np_sources)
 print(nb_data_cols)
 
 combined_dataframe=pandas.DataFrame()
+
+giant_matrix = np.zeros((N_DETECTIONS, N_SOURCES*N_COLS))
+
+index = 0
 
 for file in os.listdir(MATRIX_LOCATION):
     if file.endswith(".csv"):
@@ -55,6 +80,10 @@ for file in os.listdir(MATRIX_LOCATION):
         date = split_name[1]
         time = split_name[2]
 
+        #consistent index, starting at zero
+        dayofyear = datetime.datetime.strptime(date, '%Y%m%d').timetuple().tm_yday - 1
+        print(dayofyear)
+
         # print("matrix: ", np_matrix.shape)
 
 
@@ -65,13 +94,36 @@ for file in os.listdir(MATRIX_LOCATION):
         # print()
         results[index]=result
 
+
[…]

### 2022-09-20 notes/notes.tex
optimisation problem

@@
 \documentclass{article}
-\usepackage{amsmath, amssymb, dsfont, physics, hyperref, cleveref}
+\usepackage{amsmath, amssymb, amsthm, dsfont, physics, hyperref, cleveref}
 \usepackage[margin=1.5cm]{geometry}
@@
 \date{}
+
+\newtheorem{proposition}{Proposition}
 
@@
 In this expression, the unknown is $W$ and should be fit to the data. 
-There are a plethora of numerical techniques for solving linear equations where the unknown is a matrix.
-In the following, we provide a theoretical description of a system that can be solved for $W$.
+% There are a plethora of numerical techniques for solving linear equations where the unknown is a matrix.
+% In the following, we provide a theoretical description of a system that can be solved for $W$.
 
-We can put constraints on $W$ of the form $W \in \mathcal{W}$, where $\mathcal{W}$ is some set of constrained diagonal matrices.
+% We can put constraints on $W$ of the form $W \in \mathcal{W}$, where $\mathcal{W}$ is some set of constrained diagonal matrices.
 Suppose we have a parametrisation $\Gamma: \mathbb{R}^{k} \rightarrow \mathcal{W}, u \mapsto \Gamma(u)$.
-In this case, \cref{eq: time-evolving model concise} can be reformulated as a linear operation applied to $\Gamma(u)$ as follows:
+% In this case, \cref{eq: time-evolving model concise} can be reformulated as a linear operation applied to $\Gamma(u)$ as follows:
+% $$
+% \hat{y} = ((S^{-1} x \otimes \vb{1}_{365})^T \otimes M_{shift}) \mathrm{vec}\, \Gamma(u)
+% .$$
+% If $\Gamma$ is a linear map, the above is a linear expression in $u$. Then, the minimisation of $\norm{y - \hat{y}}^2$ can be solved using numerical linear algebra.
+
+\subsection{Fitting the model}
+The indices $i = 1,\dots,n$ are partitioned as a disjoint union $\mathcal{I}_{det} \cup \mathcal{I}_{0}$ where $\mathcal{I}_0$ contains the indices of non-detection events.
+Let $P_{det}$ be the linear map that removes the rows of a matrix that correspond to $\mathcal{I}_0$.
+We can measure the accuracy as $\norm{\log.(P_{det} \hat{y})}^2$ where $\log.(\cdot)$ is the elementwise natural logarithm.
+The simplicity of $W$ can be measured as $\norm{\log W}^2$.
+If we forget about non-detections, we have the cost function
 $$
-\hat{y} = ((S^{-1} x \otimes \vb{1}_{365})^T \otimes M_{shift}) \mathrm{vec}\, \Gamma(u)
+\min_{W \text{ diagonal}} f(W) =
+\frac{1}{2}\norm{F(W)}^2 + \frac{1}{2}\lambda \norm{\log W}^2
 .$$
-If $\Gamma$ is a linear map, the above is a linear expression in $u$. Then, the minimisation of $\norm{y - \hat{y}}^2$ can be solved using numerical linear algebra.
+\begin{proposition}
+    Let $z$ be the vector obtained by applying the function $x \mapsto (\log x) / x$ to all predicted concentrations $\hat{y}_i$ corresponding to detection events.
+Then the gradient of $f(W)$ is the diagonal matrix
+    $$
+    \mathrm{grad}_{W} f(W) = \mathrm{Diag}\left((P_{det} M_{shift})^T z \tilde{x}^T\right) + \lambda W^{-1} \log W
+    $$
+    where $\mathrm{Diag}$ is the map that sets the off-diagonal elements of a matrix to zero.
+\end{proposition}
+\begin{proof}[Proof (still to be checked for calculation errors)]
+
+Parameterise $W$ linearly as $\mathrm{vec}\, W(t) = U t$ for some unitary $U$.
+Define $$
+G(t) := 
+P_{det} M_{shift} W(t) \tilde{x}_{365}
+\quad\text{and}\quad
+F(t) := \log.(G(t))
+.$$
+Then 
+$$
+G(t) = \underbrace{P_{det} M_{shift} W(t) \tilde{x}_{365} = (P_{det}M_{shift} \oslash \tilde{x}^T) U}_A t
+$$
+where $\oslash$ is the Kronecker product with reversed arguments, which satisfies $\mathrm{vec}(A B C) = (A \oslash C^T) \mathrm{vec} B$.
+Since $G(t)$ is linear, $F(t) = \log.(G(t))$ is easy to differentiate:
+$$
+\frac{\partial}{\partial t} F(t) =
+\frac{\partial}{\partial t} \log.(G(t)) = \mathrm{diag}\,(G(t))^{-1} A
+.$$
+Using standard differentiation rules for the squared norm of a vector-valued function, we obtain
+\begin{align*}
+\mathrm{grad}_t\, \left(\frac{1}{2} \norm{F(t)}^2 \right) = \left(\frac{\partial F}{\partial t}\right)^T F(t)
+&= A^T \mathrm{diag}(G(t))^{-1} \log.(G(t)) \\
+&= A^T z \\
+&= U^T ((P_{det} M_{shift})^T \oslash \tilde{x}) z \\
+&= U^T (P_{det} M_{shift})^T z \tilde{x}^T
+.
+\end{align*}
+Since $U$ is unitary, we have $$
+\mathrm{grad}_W \left(\frac{1}{2} \norm{F(t)}^2 \right)
+=
+U \mathrm{grad}_t\, \left(\frac{1}{2} \norm{F(t)}^2 \right)
+=
+UU^T (P_{det} M_{shift})^T z \tilde{x}^T
+.$$
+The linear map $UU^T$ is simply the orthogonal projector onto the diagonal matrices, which is equivalent to $\mathrm{Diag}$.
+
+The foregoing gives the derivative of the first term in $f(W)$. The gradient of the second term can be calculated in exactly the same manner: 
+$$
+\mathrm{grad}_W\, \left(\frac{1}{2}\norm{\log W}^2\right) = W^{-1} \log W
+.$$
+\end{proof}
 

### 2022-09-20 code/import.py
Added simple least squares computation of the sources. Note: the actual computation will be wrong due to the problem being very underdetermined

@@ -2,7 +2,7 @@ import pandas
 import os
 import numpy as np
 import matplotlib.pyplot as plt
-import scipy.sparse as sp
+import scipy.optimize as opt
 import datetime
 
 MATRIX_LOCATION = "./../data/srs_csv_files/"
@@ -131,12 +131,44 @@ print("size giant matrix: ", giant_matrix.shape)
 #we do not actually want to deal with the negative values, nor with rows of only zeros (as some gap in the data exists temporally)
 pruned_giant_matrix = giant_matrix[(giant_matrix.min(axis=1)>=0.0) & (giant_matrix.max(axis=1)>0.0)]
 print("size pruned matrix: ", pruned_giant_matrix.shape)
+print("nb cols pruned matrix: ", pruned_giant_matrix.shape[0])
+print("nb zero cols pruned matrix: ", pruned_giant_matrix[:, (pruned_giant_matrix.max(axis=0)==0.0)].shape)
+#also stripping away all zero columns?, as they cannot be computed
+#FIXME: add some way of mapping these indicies to the original indices
+# pruned_giant_matrix = pruned_giant_matrix[:, (pruned_giant_matrix.max(axis=0)>0.0)]
+nb_rows_pruned_matrix = pruned_giant_matrix.shape[0]
+nb_cols_pruned_matrix = pruned_giant_matrix.shape[1]
+print("nb cols pruned matrix: ", nb_cols_pruned_matrix)
+_, singular_values, _ =np.linalg.svd(pruned_giant_matrix)
+print
+
 #TODO: prune zero columns?, as these cannot be determineds
 print(scalings)
 print(sources)
+
+
+
+def ln_y_squared(xval_array):
+    lny = np.log(pruned_giant_matrix @ xval_array)
+    return lny.T @ lny
 
 plt.hist(np.log10(results), bins=50)
 plt.xlabel("log10(y_sim)")
 plt.legend()
 plt.title("Simulated observations")
+# plt.show()
+
+print(ln_y_squared(np.ones((nb_cols_pruned_matrix,1))))
+
+#just least squares without any constraints
+# print(np.ones((nb_cols_pruned_matrix,1)).shape)
+solx=np.linalg.lstsq(pruned_giant_matrix, np.ones((nb_rows_pruned_matrix,1)))[0]
+print(solx)
+plt.figure(1)
+plt.plot(solx[:,0])
 plt.show()
+
+#initial guess is evidently the starting data for x
[…]

### 2022-09-20 notes/notes.tex
update optimisation formulation
fix typo in notation

@@
 \min_{W \text{ diagonal}} f(W) =
-\frac{1}{2}\norm{F(W)}^2 + \frac{1}{2}\lambda \norm{\log W}^2
+\frac{1}{2}\norm{\log.(P_{det} M_{shift} W(t) \tilde{x}_{365})}^2 + \frac{1}{2}\lambda \norm{\log W}^2
 .$$
+The first term measures the average logarithm of the predicted values. The second term measures the average logarithm of the weights.
 \begin{proposition}
@@
     $$
-    \mathrm{grad}_{W} f(W) = \mathrm{Diag}\left((P_{det} M_{shift})^T z \tilde{x}^T\right) + \lambda W^{-1} \log W
+    \mathrm{grad}_{W} f(W) = \mathrm{Diag}\left((P_{det} M_{shift})^T z \tilde{x}_{365}^T\right) + \lambda W^{-1} \log W
     $$
@@
 $$
-G(t) = \underbrace{P_{det} M_{shift} W(t) \tilde{x}_{365} = (P_{det}M_{shift} \oslash \tilde{x}^T) U}_A t
+G(t) = \underbrace{P_{det} M_{shift} W(t) \tilde{x}_{365} = (P_{det}M_{shift} \oslash \tilde{x}_{365}^T) U}_A t
 $$
@@
 &= A^T z \\
-&= U^T ((P_{det} M_{shift})^T \oslash \tilde{x}) z \\
-&= U^T (P_{det} M_{shift})^T z \tilde{x}^T
+&= U^T ((P_{det} M_{shift})^T \oslash \tilde{x}_{365}) z \\
+&= U^T (P_{det} M_{shift})^T z \tilde{x}_{365}^T
 .
@@
 =
-UU^T (P_{det} M_{shift})^T z \tilde{x}^T
+UU^T (P_{det} M_{shift})^T z \tilde{x}_{365}^T
 .$$

### 2022-09-20 code/import.py
Metrics to evaluate fit of current model

@@ -82,3 +82,9 @@ plt.legend()
 plt.title("Simulated observations")
 plt.show()
 
+# Caculate metrics
+MB = np.mean(results[np.where(results>=0)])-1 # mean Model Bias
+FB = 2*((1-np.mean(results[np.where(results>=0)]))/(1+np.mean(results[np.where(results>=0)]))) # Fractional Bias
+MG = np.exp(-np.mean(np.log(results[np.where(results>=0)]))) # Geometric Mean Bias
+FAC2 = np.size(np.where((results < 2) & (results > .5)))/np.size(np.where(results>=0)) # Fraction of the data where the model is within a 
+#log10Error = np.mean(np.log10(results[np.where(results>=0)]))

### 2022-09-20 notes/notes.tex
simplify equation

@@
 Let $P_{det}$ be the linear map that removes the rows of a matrix that correspond to $\mathcal{I}_0$.
-We can measure the accuracy as $\norm{\log.(P_{det} \hat{y})}^2$ where $\log.(\cdot)$ is the elementwise natural logarithm.
+We can measure the accuracy as $\norm{\log.(P_{det} \hat{y})}^2$ where $\log.(\cdot)$ is the elementwise \emph{natural} logarithm.
 The simplicity of $W$ can be measured as $\norm{\log W}^2$.
@@
 \min_{W \text{ diagonal}} f(W) =
-\frac{1}{2}\norm{\log.(P_{det} M_{shift} W(t) \tilde{x}_{365})}^2 + \frac{1}{2}\lambda \norm{\log W}^2
+\frac{1}{2}\norm{\log.(P_{det} M_{shift} W \tilde{x}_{365})}^2 + \frac{1}{2}\lambda \norm{\log W}^2
 .$$
-The first term measures the average logarithm of the predicted values. The second term measures the average logarithm of the weights.
+The first term measures the average logarithm of the predicted values $\hat{y}$ (detection events only). The second term measures the average logarithm of the weights.
 \begin{proposition}
-    Let $z$ be the vector obtained by applying the function $x \mapsto (\log x) / x$ to all predicted concentrations $\hat{y}_i$ corresponding to detection events.
+    Let $\hat{z}$ be the vector $[\hat{z}_1\,\,\dots\,\,\hat{z}_n]^T$ where 
+    $$
+    \hat{z}_i = \begin{cases}
+        \frac{\log \hat{y}_i}{\hat{y}_i} & \text{if observation $i$ is a detection} \\
+        0 & \text{if observation $i$ is a non-detection} 
+    \end{cases}
+    .$$
     Then the gradient of $f(W)$ is the diagonal matrix
     $$
-    \mathrm{grad}_{W} f(W) = \mathrm{Diag}\left((P_{det} M_{shift})^T z \tilde{x}_{365}^T\right) + \lambda W^{-1} \log W
+    \mathrm{grad}_{W} f(W) = \mathrm{Diag}\left(M_{shift}^T \hat{z} \tilde{x}_{365}^T\right) + \lambda W^{-1} \log W
     $$
@@
 \end{proposition}
+
 \begin{proof}[Proof (still to be checked for calculation errors)]
@@
 G(t) :=
+P_{det} \hat{y}_i
+=
 P_{det} M_{shift} W(t) \tilde{x}_{365}
@@
 .$$
+Write $z := P_{det} \hat{z}$.
+We can interpret $z$ as the vector obtained by applying the map $x \mapsto (\log x) / x$ elementwise to $G(t)$.
 Using standard differentiation rules for the squared norm of a vector-valued function, we obtain
@@
 &= U^T ((P_{det} M_{shift})^T \oslash \tilde{x}_{365}) z \\
-&= U^T (P_{det} M_{shift})^T z \tilde{x}_{365}^T
-.
+&= U^T (P_{det} M_{shift})^T z \tilde{x}_{365}^T \\
+&= U^T M_{shift}^T P_{det}^T P_{det} \hat{z} \tilde{x}_{365}^T \\
+&= U^T M_{shift}^T z \tilde{x}_{365}^T
 \end{align*}
+where the last equality follows from the fact that $P_{det}^T P_{det}$ is the map that sets the rows corresponding to the non-detection events to zero. 
 Since $U$ is unitary, we have $$
@@
 UU^T (P_{det} M_{shift})^T z \tilde{x}_{365}^T
-.$$
+$$
 The linear map $UU^T$ is simply the orthogonal projector onto the diagonal matrices, which is equivalent to $\mathrm{Diag}$.

### 2022-09-21 notes/notes.tex
update notation

@@
 Now we present a different interpretation.
-Instead of $S^{-1} x \otimes \vb{1}_{15}$, we make a vector $\tilde{x}_{365} := S^{-1} x \otimes \vb{1}_{365}$.
+Instead of $S^{-1} x \otimes \vb{1}_{15}$, we make a vector $\tilde{x} := S^{-1} x \otimes \vb{1}_{365}$.
 This is a vector of length $365 \times 200$ that gives the emission from day 1 to day 365 for each source.
 This is assumed to be constant for all days.
-Therefore, $\tilde{x}_{365}$ consists of $m = 200$ blocks, each of which is a constant vector of length $365$.
+Therefore, $\tilde{x}$ consists of $m = 200$ blocks, each of which is a constant vector of length $365$.
 
-We can write a system of equations that is logically the same as \cref{eq: initial model concise} but uses $\tilde{x}_{365}$.
+We can write a system of equations that is logically the same as \cref{eq: initial model concise} but uses $\tilde{x}$.
 This gives a larger linear system.
@@
 }_{ := M_{shift}}
-\tilde{x}_{365}
+\tilde{x}
 \end{equation}
@@
 $$
-\tilde{x}_{365} := S^{-1} x \otimes \vb{1}_{365}
+\tilde{x} := S^{-1} x \otimes \vb{1}_{365}
 =
@@
 
-We could reorganise the vector $\tilde{x}_{365}$ to the transposed expression $\vb{1}_{365} \otimes (S^{-1} x)$, which is a vector containing all emissions on day 1, followed by all emissions on day 2, etc.
+We could reorganise the vector $\tilde{x}$ to the transposed expression $\vb{1}_{365} \otimes (S^{-1} x)$, which is a vector containing all emissions on day 1, followed by all emissions on day 2, etc.
 A similar permutation of $y$ could be chosen.
@@
 .$$
-Using this weighted vector of emissions instead of $\tilde{x}_{365}$ in \cref{eq: initial model 365 days} gives
+Using this weighted vector of emissions instead of $\tilde{x}$ in \cref{eq: initial model 365 days} gives
 \begin{equation}
@@
 \min_{W \text{ diagonal}} f(W) =
-\frac{1}{2}\norm{\log.(P_{det} M_{shift} W \tilde{x}_{365})}^2 + \frac{1}{2}\lambda \norm{\log W}^2
+\underbrace{\frac{1}{2}\norm{\log.(P_{det} M_{shift} W \tilde{x})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{\log W}^2}_{\text{deviation from average emission}}
 .$$
-The first term measures the average logarithm of the predicted values $\hat{y}$ (detection events only). The second term measures the average logarithm of the weights.
 \begin{proposition}
@@
     $$
-    \mathrm{grad}_{W} f(W) = \mathrm{Diag}\left(M_{shift}^T \hat{z} \tilde{x}_{365}^T\right) + \lambda W^{-1} \log W
+    \mathrm{grad}_{W} f(W) = \mathrm{Diag}\left(M_{shift}^T \hat{z} \tilde{x}^T\right) + \lambda W^{-1} \log W
     $$
@@
 =
-P_{det} M_{shift} W(t) \tilde{x}_{365}
+P_{det} M_{shift} W(t) \tilde{x}
 \quad\text{and}\quad
@@
 $$
-G(t) = \underbrace{P_{det} M_{shift} W(t) \tilde{x}_{365} = (P_{det}M_{shift} \oslash \tilde{x}_{365}^T) U}_A t
+G(t) = \underbrace{P_{det} M_{shift} W(t) \tilde{x} = (P_{det}M_{shift} \oslash \tilde{x}^T) U}_A t
 $$
@@
 &= A^T z \\
-&= U^T ((P_{det} M_{shift})^T \oslash \tilde{x}_{365}) z \\
-&= U^T (P_{det} M_{shift})^T z \tilde{x}_{365}^T \\
-&= U^T M_{shift}^T P_{det}^T P_{det} \hat{z} \tilde{x}_{365}^T \\
-&= U^T M_{shift}^T z \tilde{x}_{365}^T
+&= U^T ((P_{det} M_{shift})^T \oslash \tilde{x}) z \\
+&= U^T (P_{det} M_{shift})^T z \tilde{x}^T \\
+&= U^T M_{shift}^T P_{det}^T P_{det} \hat{z} \tilde{x}^T \\
+&= U^T M_{shift}^T z \tilde{x}^T
 \end{align*}
@@
 =
-UU^T (P_{det} M_{shift})^T z \tilde{x}_{365}^T
+UU^T (P_{det} M_{shift})^T z \tilde{x}^T
 $$

### 2022-09-21 code/constants.py
Splitted file for code clarity.

@@ -1,0 +1,6 @@
+#Just a file defining some constants for these scripts
+LIMIT_BACKWARD_TIME = 15
+N_SOURCES = 200
+N_DETECTIONS = 4636
+#allowing us to use values for 1 entire year
+N_COLS=365+LIMIT_BACKWARD_TIME

### 2022-09-21 code/import.py

@@ -1,180 +1,0 @@
-import pandas
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import scipy.optimize as opt
-import datetime
-
-MATRIX_LOCATION = "./../data/srs_csv_files/"
-#TODO: figure out these parameters from the files themselves
-LIMIT_BACKWARD_TIME = 15
-N_SOURCES = 200
-N_DETECTIONS = 4636
-#allowing us to use values for 1 entire year
-N_COLS=365+LIMIT_BACKWARD_TIME
-
-
-scalings=pandas.read_csv("./../data/scalings.csv", header=None)
-sources=pandas.read_csv("./../data/data_xe133_NH_yearly.csv")#has a header file, not that we actually need it
-
-print(scalings)
-print(sources)
-
-np_scalings=scalings[0].to_numpy()
-np_sources=sources['emission'].to_numpy()*1000#emission in mBq instead of Bq
-# print(np_sources.size)
-# print(np_sources.shape)
-#rescaling sources
-# print("scalings: ", np_scalings)
-# print("before dividing: ", np_sources)
-np_sources=np_sources/np_scalings
-# print("after dividing:", np_sources)
-#np.divide(np_sources,np_scalings)
-
-
-# print(np_scalings.size)
-# print(np_scalings.shape)
-# print("np sources: ", np_sources.size)
-
-n_datafiles=0
-for file in os.listdir(MATRIX_LOCATION):
-    if file.endswith(".csv"):
-        n_datafiles+=1
-results=np.zeros(n_datafiles)
-print(results)
-
-
-
-
-#TODO figure out whether this thing needs to be sparse
-#probably
-# def fill_matrix(y, index, ):
-    # row =
-    # col =
-    # data =
-#or just construct from standard numpy array...
-
-
-
-index=0
-
-#timesteps times number of emissions
-nb_data_cols=np.size(np_sources)
-print(nb_data_cols)
-
-combined_dataframe=pandas.DataFrame()
-giant_matrix = np.zeros((N_DETECTIONS, N_SOURCES*N_COLS))
-measurements_dict = {};
-
-
-
-for file in os.listdir(MATRIX_LOCATION):
-    if file.endswith(".csv"):
-        matrix_file = pandas.read_csv(os.path.join(MATRIX_LOCATION, file), header=None)
-        np_matrix = matrix_file.to_numpy()
-        #prune extension from filename
-        pruned_name = file.split('.')[0]
-        split_name = pruned_name.split('_')
[…]

### 2022-09-21 code/main.py

@@ -1,0 +1,23 @@
+import readin as read
+import analysis as analysis
+import numpy as np
+import constants as cst
+import matplotlib.pyplot as plt
+
+np_sources = read.import_scaled_sources()
+print("importing matrix")
+#import the combined matrix
+giant_matrix = read.import_combined_matrix()
+print("done importing matrix")
+#and throw away all zero/negative rows
+pruned_giant_matrix = read.prune_matrix(giant_matrix)
+
+#default source assumes all sources to be constant at all times
+default_x = np.kron(np_sources, np.ones(cst.N_COLS))
+y_sim = analysis.evaluate_x(pruned_giant_matrix, default_x)
+
+plt.hist(np.log10(y_sim), bins=50)
+plt.xlabel("log10(y_sim)")
+plt.legend()
+plt.title("Simulated observations")
+plt.show()

### 2022-09-21 code/readin.py

@@ -1,0 +1,243 @@
+import pandas
+import os
+import numpy as np
+import scipy.optimize as opt
+import datetime
+import constants as cst
+
+MATRIX_LOCATION = "./../data/srs_csv_files/"
+# #TODO: figure out these parameters from the files themselves?
+# LIMIT_BACKWARD_TIME = 15
+# N_SOURCES = 200
+# N_DETECTIONS = 4636
+# #allowing us to use values for 1 entire year
+# N_COLS=365+LIMIT_BACKWARD_TIME
+
+
+# scalings=pandas.read_csv("./../data/scalings.csv", header=None)
+# sources=pandas.read_csv("./../data/data_xe133_NH_yearly.csv")#has a header file, not that we actually need it
+#
+# print(scalings)
+# print(sources)
+#
+# np_scalings=scalings[0].to_numpy()
+# np_sources=sources['emission'].to_numpy()*1000#emission in mBq instead of Bq
+# # print(np_sources.size)
+# # print(np_sources.shape)
+# #rescaling sources
+# # print("scalings: ", np_scalings)
+# # print("before dividing: ", np_sources)
+# np_sources=np_sources/np_scalings
+# # print("after dividing:", np_sources)
+# #np.divide(np_sources,np_scalings)
+#
+#
+# # print(np_scalings.size)
+# # print(np_scalings.shape)
+# # print("np sources: ", np_sources.size)
+#
+# n_datafiles=0
+# for file in os.listdir(MATRIX_LOCATION):
+#     if file.endswith(".csv"):
+#         n_datafiles+=1
+# results=np.zeros(n_datafiles)
+# print(results)
+
+
+
+
+#TODO figure out whether this thing needs to be sparse
+#probably
+# def fill_matrix(y, index, ):
+    # row =
+    # col =
+    # data =
+#or just construct from standard numpy array...
+
+def import_scalings():
+    scalings=pandas.read_csv("./../data/scalings.csv", header=None)
+    np_scalings=scalings[0].to_numpy()
+    return np_scalings
+
+def import_sources():
+    sources=pandas.read_csv("./../data/data_xe133_NH_yearly.csv")#has a header file, not that we actually need it
+    np_sources=sources['emission'].to_numpy()*1000#emission in mBq instead of Bq
+    return np_sources
+
+def import_scaled_sources():
+    return import_sources()/import_scalings()
+
+# index=0
+
+# #timesteps times number of emissions
+# nb_data_cols=np.size(np_sources)
+# print(nb_data_cols)
+
[…]

### 2022-09-21 notes/notes.tex
numerical check of gradient

@@
 \end{proposition}
-
-\begin{proof}[Proof (still to be checked for calculation errors)]
+\begin{proof}
 
@@
 $$
-G(t) = \underbrace{P_{det} M_{shift} W(t) \tilde{x} = (P_{det}M_{shift} \oslash \tilde{x}^T) U}_A t
-$$
-where $\oslash$ is the Kronecker product with reversed arguments, which satisfies $\mathrm{vec}(A B C) = (A \oslash C^T) \mathrm{vec} B$.
+G(t) = P_{det} M_{shift} W(t) \tilde{x} = \underbrace{(\tilde{x}^T \otimes P_{det}M_{shift}) U}_A t
+.$$
 Since $G(t)$ is linear, $F(t) = \log.(G(t))$ is easy to differentiate:
@@
 &= A^T z \\
-&= U^T ((P_{det} M_{shift})^T \oslash \tilde{x}) z \\
+&= U^T (\tilde{x} \otimes (P_{det} M_{shift})^T) z \\
 &= U^T (P_{det} M_{shift})^T z \tilde{x}^T \\
@@
 \end{proof}
+For even more evidence, there is a Julia script in the repository that checks this formula against the result obtained from automatic differentiation (using simplified mock data).
+The result is correct up to machine precision.
 

### 2022-09-21 test/test_gradient_formula.jl

@@ -1,0 +1,27 @@
+using LinearAlgebra, Zygote
+
+# Generate simplified version of the data
+Iᵈᵉᵗ = [1,3,4] # indices corresponding to detection events
+Mₛ = abs.(randn(4, 10)) # M_shift (structure unimportant)
+P  = Matrix{Bool}(I, (4,4))[Iᵈᵉᵗ, :] 
+x̃  = exp.(randn(size(Mₛ,2)))
+λ  = 0.1
+
+norm²(x) = dot(x,x) # automatically differentiable
+f(W :: AbstractMatrix) = norm²(log.(P*Mₛ*W*x̃))/2 + λ*norm²(log.(diag(W)))/2 # function of diagonal matrix W
+f(w :: AbstractVector) = f(Diagonal(w)) # function of vector of diagonal elements
+
+# compute gradient using automatic differentiation
+w₀ = exp.(randn(size(x̃)))
+W₀ = Diagonal(w₀)
+∇fᴬᴰ = Diagonal(Zygote.gradient(f, w₀)[1])
+
+# gradient obtained from first principles
+ŷ = Mₛ*W₀*x̃
+ẑ = [ i ∈ Iᵈᵉᵗ ? log(ŷ[i]) / ŷ[i] : 0.0 for i ∈ axes(Mₛ,1)]
+∇f = Diagonal(Mₛ'*ẑ*x̃') + λ * (W₀ \ log(W₀))
+
+println("|| ∇f || = ", norm(∇f))
+println("|| ∇fᴬᴰ || = ", norm(∇fᴬᴰ))
+println("|| ∇f - ∇fᴬᴰ || / || ∇f || = ", norm(∇f - ∇fᴬᴰ) / norm(∇fᴬᴰ))
+

### 2022-09-21 read_geopandas.py
geo map of source contribution
WIP

@@ -1,0 +1,90 @@
+import pandas as pd
+import numpy as np
+import os
+
+# https://stackoverflow.com/questions/53233228/plot-latitude-longitude-from-csv-in-python-3-6
+# https://geopandas.org/en/latest/gallery/cartopy_convert.html
+# https://stackoverflow.com/questions/55646598/polar-stereographic-projection-of-geopandas-world-map
+# https://geopandas.org/en/latest/gallery/cartopy_convert.html
+from matplotlib import pyplot as plt
+import matplotlib.cm as cm
+from shapely.geometry import Point, Polygon, MultiPoint
+import geopandas as gpd
+import cartopy.crs as ccrs
+from shapely.ops import triangulate, voronoi_diagram
+
+m_paths = os.path.join('data', 'srs_csv_files')
+x_path = os.path.join('data', 'data_xe133_NH_yearly.csv')
+y_path = os.path.join('data', 'scalings.csv')
+
+scalings = pd.read_csv(y_path, header = None)
+sources = pd.read_csv(x_path)
+sources['scaled'] = sources['emission'] * 1000 / scalings[0]
+
+# remove duplicate days
+samples = pd.DataFrame([f.split('_') for f in os.listdir(m_paths)],
+                       columns = ['obs', 'd', 't'])
+samples.drop_duplicates(subset = ['obs', 'd'], inplace = True)
+
+# keep days with all stations
+days = samples.groupby('d')['obs'].apply(set)
+s, c = np.unique(days.values, return_counts = True)
+print(c.max(), s[c.argmax()])
+best = days[days == s[c.argmax()]]
+samples2 = samples[samples['d'].isin(best.index.values)].copy()
+
+## todo, most consecutive
+
+for row in samples2.iterrows():
+    f = '_'.join(samples2.iloc[0])
+    data = pd.read_csv(os.path.join(m_paths, f), header = None)
+    # np.matmul(data.T, sources['scaled'].values)
+    z = data.sum(axis = 1) # total over 15 days
+
+    zm = z.max() - 2 * z.median()
+    import matplotlib as mpl
+    norm = mpl.colors.Normalize(vmin = z.median() - 2 * zm, vmax = z.median() + 2 * zm)
+    cmap = cm.hot
+    m = cm.ScalarMappable(norm = norm, cmap = cmap)
+    colos = m.to_rgba(z)
[…]

### 2022-09-21 code/readin.py
Corrected position of vectors in matrix

@@ -92,7 +92,7 @@ def import_combined_matrix():
             time = split_name[2]
 
             #consistent index, starting at zero
-            dayofyear = datetime.datetime.strptime(date, '%Y%m%d').timetuple().tm_yday - 1
+            # dayofyear = datetime.datetime.strptime(date, '%Y%m%d').timetuple().tm_yday - 1
             # print(dayofyear)
 
             # print("matrix: ", np_matrix.shape)
@@ -106,9 +106,12 @@ def import_combined_matrix():
         # print(dictel)
         np_matrix = val["matrix"]
 
+
         #setting up some larger matrix for eventually filling in the combined matrix
         largermat=np.zeros((cst.N_SOURCES, cst.N_COLS))
-        date=dayofyear
+        date_name = val["date"]
+        dayofyear = datetime.datetime.strptime(date_name, '%Y%m%d').timetuple().tm_yday - 1
+        date = dayofyear
 
         #and putting it in in the correct place?
         largermat[:, date:date+cst.LIMIT_BACKWARD_TIME]=np_matrix

### 2022-09-21 code/main.py
Added gradient of cost function

@@ -20,4 +20,11 @@ plt.hist(np.log10(y_sim), bins=50)
 plt.xlabel("log10(y_sim)")
 plt.legend()
 plt.title("Simulated observations")
-plt.show()
+# plt.show()
+
+nb_rows = giant_matrix.shape[1]
+print(nb_rows)
+
+# print(analysis.grad_cost_function(np.ones(nb_rows)))
+W = np.ones(default_x.shape)
+print(analysis.grad_cost_function(y_sim, default_x, pruned_giant_matrix, W, 1, 0.001))

### 2022-09-21 notes/notes.tex
simplify expression for gradient

@@
     \item $\vb{1}_q = [1\,\, \dots\,\, 1]^T \in \mathbb{R}^q$ is the constant vector consisting of only ones.
+    \item $\odot$ is the Hadamard (i.e.\, elementwise) product, given by $(AB)_{ij} = A_{ij} B_{ij}$.
+    \item $\mathrm{diag}$ is the canonical identification of $p$-vectors and $p \times p$ diagonal matrices, i.e.,
+    $$
+    \mathrm{diag}\left(
+    \begin{bmatrix}
+        a_1 \\ 
+        \vdots \\
+        a_p
+    \end{bmatrix}
+    \right)
+    =
+    \begin{bmatrix}
+        a_1 & 0 & \dots & 0 \\
+        0 & a_2 & \dots & 0 \\
+        \vdots & \vdots & \ddots & 0 \\
+        0 & 0 & \cdots & a_p
+    \end{bmatrix}
+    .$$
     \item $\otimes$ is the Kronecker product. In particular,
@@
     $$
-    \mathrm{grad}_{W} f(W) = \mathrm{Diag}\left(M_{shift}^T \hat{z} \tilde{x}^T\right) + \lambda W^{-1} \log W
-    $$
-    where $\mathrm{Diag}$ is the map that sets the off-diagonal elements of a matrix to zero.
+    \mathrm{grad}_{W} f(W) = \mathrm{diag}\left(\left(M_{shift}^T \hat{z}\right) \odot \tilde{x}\right) + \lambda W^{-1} \log W
+    .$$
 \end{proposition}
@@
 &= U^T M_{shift}^T P_{det}^T P_{det} \hat{z} \tilde{x}^T \\
-&= U^T M_{shift}^T z \tilde{x}^T
+&= U^T M_{shift}^T \hat{z} \tilde{x}^T
 \end{align*}
@@
 =
-UU^T (P_{det} M_{shift})^T z \tilde{x}^T
+UU^T M_{shift}^T \hat{z} \tilde{x}^T
 $$
-The linear map $UU^T$ is simply the orthogonal projector onto the diagonal matrices, which is equivalent to $\mathrm{Diag}$.
+The linear map $UU^T$ is simply the orthogonal projector onto the diagonal matrices.
+This orthogonal projection simply sets the off-diagonal elements of its argument to zero.
+Since the diagonal of $M_{shift}^T \hat{z} \tilde{x}^T$ is $(M_{shift}^T \hat{z}) \odot \tilde{x}$, we have
+$$
+\mathrm{grad}_W \left(\frac{1}{2} \norm{F(t)}^2 \right)
+=
+\mathrm{diag}((M_{shift}^T \hat{z}) \odot \tilde{x})
+.$$
 

### 2022-09-21 test/test_gradient_formula.jl

@@ -19,7 +19,7 @@ W₀ = Diagonal(w₀)
 # gradient obtained from first principles
 ŷ = Mₛ*W₀*x̃
 ẑ = [ i ∈ Iᵈᵉᵗ ? log(ŷ[i]) / ŷ[i] : 0.0 for i ∈ axes(Mₛ,1)]
-∇f = Diagonal(Mₛ'*ẑ*x̃') + λ * (W₀ \ log(W₀))
+∇f = Diagonal((Mₛ'*ẑ) .* x̃) + λ * (W₀ \ log(W₀))
 
 println("|| ∇f || = ", norm(∇f))
 println("|| ∇fᴬᴰ || = ", norm(∇fᴬᴰ))

### 2022-09-21 code/analysis.py
First attempt at gradient descent

@@ -1,35 +1,64 @@
 import numpy as np
+import constants as cst
+import scipy as sp
 
 def ln_y_squared(giant_matrix, xval_array):
     lny = np.log(giant_matrix @ xval_array)
     return lny.T @ lny
 
-#W is vector of W
-def cost_function(giant_matrix, W, reg_lambda):
+#W is vector
+#TODO FIXME
+#xval_array = W * default_x
+def cost_function(giant_matrix, xval_array, W, reg_lambda):
     # TODO: use W instead of xval_array; hint: reconstruct xvalues using this W (just kron as below)
     lny = np.log(giant_matrix @ xval_array)
-    regularization = reg_lambda * np.log(weigths)
+    regularization = reg_lambda * np.log(W)
+    print("cost: ", 1.0/2.0*(lny.T @ lny +regularization.T @ regularization))
+    print("min weight: ", np.min(W))
     return 1.0/2.0*(lny.T @ lny +regularization.T @ regularization)
 
-def evaluate_x(giant_matrix, xval_array):
+def evaluate_y(giant_matrix, xval_array):
     return giant_matrix @ xval_array
 
 
 
 #Gradient descent; is returned as a vector instead of a diagonal matrix
 #W is assumed to be a vector (corresponds with a diagonal matrix)
-def grad_cost_function(y, x_default, giant_matrix, W, reg_lambda, stepsize):
-    print("matrix shape_T: ", giant_matrix.T.shape)
-    print("y shape: ", y.shape)
-    print("matrix_T times y shape: ", (giant_matrix.T @ (np.log(y)/y)).shape)
-    print("x_T shape: ", x_default.T.shape)
+def grad_cost_function(y, x_default, giant_matrix, W, reg_lambda):
+    # print("matrix shape_T: ", giant_matrix.T.shape)
+    # print("y shape: ", y.shape)
+    # print("matrix_T times y shape: ", (giant_matrix.T @ (np.log(y)/y)).shape)
+    # print("x_T shape: ", x_default.T.shape)
     #to determine the diagonal of the large matrix, we do not explicitly need to compute this large matrix
     # matrix_cost = np.diag((giant_matrix.T @ (np.log(y)/y)) @ x_default.T)
     #this 'diagonal matrix' is stored in vector format
     matrix_cost = (giant_matrix.T @ (np.log(y)/y)) * x_default
     reg_cost = reg_lambda / W * np.log(W)
[…]

### 2022-09-21 code/main.py

@@ -14,7 +14,7 @@ pruned_giant_matrix = read.prune_matrix(giant_matrix)
 
 #default source assumes all sources to be constant at all times
 default_x = np.kron(np_sources, np.ones(cst.N_COLS))
-y_sim = analysis.evaluate_x(pruned_giant_matrix, default_x)
+y_sim = analysis.evaluate_y(pruned_giant_matrix, default_x)
 
 plt.hist(np.log10(y_sim), bins=50)
 plt.xlabel("log10(y_sim)")
@@ -27,4 +27,19 @@ print(nb_rows)
 
 # print(analysis.grad_cost_function(np.ones(nb_rows)))
 W = np.ones(default_x.shape)
-print(analysis.grad_cost_function(y_sim, default_x, pruned_giant_matrix, W, 1, 0.001))
+print(analysis.grad_cost_function(y_sim, default_x, pruned_giant_matrix, W, 1))
+
+
+#just some random value to try out gradient descent
+reg_lambda = 8.0
+# reg_lambda = 1.0
+start_W = np.ones(default_x.shape)
+stepsize = 0.05
+
+
+grad = analysis.grad_cost_function(y_sim, default_x, pruned_giant_matrix, W, reg_lambda)
+
+resulting_W = analysis.gradient_descent_algorithm(reg_lambda, np_sources, pruned_giant_matrix, start_W)
+plt.figure()
+plt.hist(resulting_W, bins=50)
+plt.show()

### 2022-09-21 code/main.py
Added option of only reading in the first N days of the year. Using the smaller matrix, we can also compute singular values.

@@ -22,8 +22,9 @@ plt.legend()
 plt.title("Simulated observations")
 # plt.show()
 
-nb_rows = giant_matrix.shape[1]
-print(nb_rows)
+# nb_rows = giant_matrix.shape[1]
+print("giant matrix shape: ", pruned_giant_matrix.shape)
+# print("nb cols: ", nb_rows)
 
 # print(analysis.grad_cost_function(np.ones(nb_rows)))
 W = np.ones(default_x.shape)
@@ -36,10 +37,12 @@ reg_lambda = 8.0
 start_W = np.ones(default_x.shape)
 stepsize = 0.05
 
+print(np.linalg.svd(pruned_giant_matrix))
 
-grad = analysis.grad_cost_function(y_sim, default_x, pruned_giant_matrix, W, reg_lambda)
 
-resulting_W = analysis.gradient_descent_algorithm(reg_lambda, np_sources, pruned_giant_matrix, start_W)
-plt.figure()
-plt.hist(resulting_W, bins=50)
-plt.show()
+# grad = analysis.grad_cost_function(y_sim, default_x, pruned_giant_matrix, W, reg_lambda)
+#
+# resulting_W = analysis.gradient_descent_algorithm(reg_lambda, np_sources, pruned_giant_matrix, start_W)
+# plt.figure()
+# plt.hist(resulting_W, bins=50)
+# plt.show()

### 2022-09-21 read_geopandas.py
Update read_geopandas.py

@@ -1,11 +1,9 @@
 import pandas as pd
 import numpy as np
 import os
+import itertools
+from collections import Counter
 
-# https://stackoverflow.com/questions/53233228/plot-latitude-longitude-from-csv-in-python-3-6
-# https://geopandas.org/en/latest/gallery/cartopy_convert.html
-# https://stackoverflow.com/questions/55646598/polar-stereographic-projection-of-geopandas-world-map
-# https://geopandas.org/en/latest/gallery/cartopy_convert.html
 from matplotlib import pyplot as plt
 import matplotlib.cm as cm
 from shapely.geometry import Point, Polygon, MultiPoint
@@ -26,16 +24,43 @@ samples = pd.DataFrame([f.split('_') for f in os.listdir(m_paths)],
                        columns = ['obs', 'd', 't'])
 samples.drop_duplicates(subset = ['obs', 'd'], inplace = True)
 
-# keep days with all stations
-days = samples.groupby('d')['obs'].apply(set)
-s, c = np.unique(days.values, return_counts = True)
-print(c.max(), s[c.argmax()])
-best = days[days == s[c.argmax()]]
-samples2 = samples[samples['d'].isin(best.index.values)].copy()
+# keep days with most common stations
+stations = samples['obs'].unique()
+days = samples.groupby('d')['obs'].apply(set) # groupby days, get set of available stations
 
-## todo, most consecutive
+comb = [set(v) for i in range(1, len(stations) + 1) for v in itertools.combinations(stations, i)] # all possible combinations
+count = [days.apply(lambda x: c.issubset(x)).sum() for c in comb] # count combination is subset
+description = pd.DataFrame({'subsets': comb,
+                            'len': [len(c) for c in comb],
+                            'count': count})
+#description.to_csv('common_subsets.csv')
 
-for row in samples2.iterrows():
+# use subset with most stations
+thresh = 100 # at least 100 days (*note later remove non consecutive)
+availb = description[description['count'] > thresh]
+availb.iloc[availb['len'].argmax()]
+subset, sublen, subcnt = availb.iloc[availb['len'].argmax()]
+
+days2 = days[days.apply(lambda x: subset.issubset(x))].copy()
[…]

### 2022-09-21 notes/notes.tex
reformulate theory using W = exp(V)

@@
 .$$
+In order to optimise $f$ numerically, we need to be able to measure changes to $W$.
+We do this implicitly by parametrising $W = e^{\mathrm{diag}\, v}$ for some vector $v$.
+If we apply an absolute change $v' := v + \Delta v$ in numerical optimisation, this corresponds to the relative change $W' = W e^{\mathrm{diag} \Delta v}$.
+Reformulating the cost function gives 
+$$
+\min_{v} f(v) =
+\underbrace{\frac{1}{2}\norm{\log.(P_{det} M_{shift} e^{\mathrm{diag}\, v} \tilde{x})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{v}^2}_{\text{deviation from average emission}}
+.$$
+
 \begin{proposition}
@@
     .$$
-    Then the gradient of $f(W)$ is the diagonal matrix
+    Then
     $$
-    \mathrm{grad}_{W} f(W) = \mathrm{diag}\left(\left(M_{shift}^T \hat{z}\right) \odot \tilde{x}\right) + \lambda W^{-1} \log W
+    \mathrm{grad}_{v} f(v) = e^v \odot (M_{shift}^T \hat{z}) \odot \tilde{x} + \lambda v
     .$$
@@
 
-Parameterise $W$ linearly as $\mathrm{vec}\, W(t) = U t$ for some unitary $U$.
+Let $U = \mathrm{vec} \circ \mathrm{diag}$. Then $\mathrm{vec}\, W(v) = U e^v$ where $e^v$ denotes elementwise exponentiation.
 Define $$
-G(t) :=
+G(v) :=
 P_{det} \hat{y}_i
 =
-P_{det} M_{shift} W(t) \tilde{x}
+P_{det} M_{shift} W(v) \tilde{x}
 \quad\text{and}\quad
-F(t) := \log.(G(t))
+F(v) := \log.(G(v))
 .$$
@@
 $$
-G(t) = P_{det} M_{shift} W(t) \tilde{x} = \underbrace{(\tilde{x}^T \otimes P_{det}M_{shift}) U}_A t
+G(v) = P_{det} M_{shift} W(v) \tilde{x}
+=
+(\tilde{x}^T \otimes P_{det} M_{shift}) \mathrm{vec} W(v)
+=
+\underbrace{(\tilde{x}^T \otimes P_{det}M_{shift}) U}_A e^v
+$$
+so that 
+$$
+\frac{\partial G}{\partial v} = A e^{\mathrm{diag}\, v}
 .$$
-Since $G(t)$ is linear, $F(t) = \log.(G(t))$ is easy to differentiate:
+Differentiating $F(v) = \log.(G(v))$ gives:
 $$
-\frac{\partial}{\partial t} F(t) =
-\frac{\partial}{\partial t} \log.(G(t)) = \mathrm{diag}\,(G(t))^{-1} A
+\frac{\partial F}{\partial v}
+=
+\frac{\partial}{\partial v} \log.(G(v))
+=
+\mathrm{diag}\,(G(v))^{-1} \frac{\partial G}{\partial v}
+=
+\mathrm{diag}\,(G(v))^{-1} A e^{\mathrm{diag}\, v }
 .$$
 Write $z := P_{det} \hat{z}$.
-We can interpret $z$ as the vector obtained by applying the map $x \mapsto (\log x) / x$ elementwise to $G(t)$.
+We can interpret $z$ as the vector obtained by applying the map $x \mapsto (\log x) / x$ elementwise to $G(v)$.
 Using standard differentiation rules for the squared norm of a vector-valued function, we obtain
 \begin{align*}
-\mathrm{grad}_t\, \left(\frac{1}{2} \norm{F(t)}^2 \right) = \left(\frac{\partial F}{\partial t}\right)^T F(t)
-&= A^T \mathrm{diag}(G(t))^{-1} \log.(G(t)) \\
-&= A^T z \\
-&= U^T (\tilde{x} \otimes (P_{det} M_{shift})^T) z \\
-&= U^T (P_{det} M_{shift})^T z \tilde{x}^T \\
-&= U^T M_{shift}^T P_{det}^T P_{det} \hat{z} \tilde{x}^T \\
-&= U^T M_{shift}^T \hat{z} \tilde{x}^T
+\mathrm{grad}_v\, \left(\frac{1}{2} \norm{F(v)}^2 \right) =\left(\frac{\partial F}{\partial v}\right)^T F(v)
+&= e^{\mathrm{diag}\, v} A^T \mathrm{diag}(G(t))^{-1} \log.(G(t)) \\
+&= e^{\mathrm{diag}\, v} A^T z \\
+&= e^{\mathrm{diag}\, v} U^T (\tilde{x} \otimes (P_{det} M_{shift})^T) z \\
+&= e^{\mathrm{diag}\, v} U^T (P_{det} M_{shift})^T z \tilde{x}^T \\
+&= e^{\mathrm{diag}\, v} U^T M_{shift}^T P_{det}^T P_{det} \hat{z} \tilde{x}^T \\
+&= e^{\mathrm{diag}\, v} U^T M_{shift}^T \hat{z} \tilde{x}^T
 \end{align*}
 where the last equality follows from the fact that $P_{det}^T P_{det}$ is the map that sets the rows corresponding to the non-detection events to zero. 
-Since $U$ is unitary, we have $$
-\mathrm{grad}_W \left(\frac{1}{2} \norm{F(t)}^2 \right)
+By construction of $U$, the map $U^T$ projects its argument orthogonally onto the diagonal matrices and gives the diagonal element.
+The orthogonal projection of a matrix onto the diagonal matrices is obtained by setting the off-diagonal elements to zero.
+The diagonal elements of $M_{shift}^T \hat{z} \tilde{x}^T$ are $(M_{shift}^T \hat{z}) \odot \tilde{x}$, so that the above simplifies to
+$$
+\mathrm{grad}_v\, \left(\frac{1}{2} \norm{F(v)}^2 \right)
 =
-U \mathrm{grad}_t\, \left(\frac{1}{2} \norm{F(t)}^2 \right)
+e^{\mathrm{diag}\, v} \left((M_{shift}^T \hat{z}) \odot \tilde{x}\right)
 =
-UU^T M_{shift}^T \hat{z} \tilde{x}^T
-$$
-The linear map $UU^T$ is simply the orthogonal projector onto the diagonal matrices.
-This orthogonal projection simply sets the off-diagonal elements of its argument to zero.
-Since the diagonal of $M_{shift}^T \hat{z} \tilde{x}^T$ is $(M_{shift}^T \hat{z}) \odot \tilde{x}$, we have
-$$
-\mathrm{grad}_W \left(\frac{1}{2} \norm{F(t)}^2 \right)
-=
-\mathrm{diag}((M_{shift}^T \hat{z}) \odot \tilde{x})
+e^v \odot (M_{shift}^T \hat{z}) \odot \tilde{x}
 .$$
 
-The foregoing gives the derivative of the first term in $f(W)$. The gradient of the second term can be calculated in exactly the same manner: 
-$$
-\mathrm{grad}_W\, \left(\frac{1}{2}\norm{\log W}^2\right) = W^{-1} \log W
-.$$
+The foregoing gives the derivative of the first term in $f(W)$. The gradient of the second term can is simply $\lambda v$.
+
 \end{proof}

### 2022-09-21 test/test_gradient_formula.jl

@@ -1,5 +1,6 @@
 using LinearAlgebra, Zygote
 
+
 # Generate simplified version of the data
 Iᵈᵉᵗ = [1,3,4] # indices corresponding to detection events
 Mₛ = abs.(randn(4, 10)) # M_shift (structure unimportant)
@@ -8,18 +9,17 @@ x̃  = exp.(randn(size(Mₛ,2)))
 λ  = 0.1
 
 norm²(x) = dot(x,x) # automatically differentiable
-f(W :: AbstractMatrix) = norm²(log.(P*Mₛ*W*x̃))/2 + λ*norm²(log.(diag(W)))/2 # function of diagonal matrix W
-f(w :: AbstractVector) = f(Diagonal(w)) # function of vector of diagonal elements
+v₀ = randn(size(x̃))
+W₀ = exp(Diagonal(v₀))
 
-# compute gradient using automatic differentiation
-w₀ = exp.(randn(size(x̃)))
-W₀ = Diagonal(w₀)
-∇fᴬᴰ = Diagonal(Zygote.gradient(f, w₀)[1])
+# gradient using automatic differentiation
+f(v) = norm²(log.(P*Mₛ*exp(Diagonal(v))*x̃))/2 + λ*norm²(v)/2	
+∇fᴬᴰ = Zygote.gradient(f, v₀)[1]
 
-# gradient obtained from first principles
+# gradient using first principles derivation
 ŷ = Mₛ*W₀*x̃
 ẑ = [ i ∈ Iᵈᵉᵗ ? log(ŷ[i]) / ŷ[i] : 0.0 for i ∈ axes(Mₛ,1)]
-∇f = Diagonal((Mₛ'*ẑ) .* x̃) + λ * (W₀ \ log(W₀))
+∇f = exp.(v₀) .* (Mₛ'*ẑ) .* x̃ + λ * v₀
 
 println("|| ∇f || = ", norm(∇f))
 println("|| ∇fᴬᴰ || = ", norm(∇fᴬᴰ))

### 2022-09-21 code/analysis.py
Now tries to optimize log(W) using gradient descent.

@@ -37,17 +37,30 @@ def grad_cost_function(y, x_default, giant_matrix, W, reg_lambda):
     return (matrix_cost+reg_cost)
 
 
-
+#returns the optimized weights
 def gradient_descent_algorithm(reg_lambda, np_sources, giant_matrix, start_W):
     default_x = np.kron(np_sources, np.ones(cst.N_COLS))
     W = start_W
 
-    lambda_cost_function = lambda W :cost_function(giant_matrix, W*default_x, W, reg_lambda)
+    #optimizing lnW makes sure that W will always be positive
+    lnW = np.log(W)
+
+    #original optimization stuff for W
+    lambda_cost_function = lambda W : cost_function(giant_matrix, W*default_x, W, reg_lambda)
     lambda_gradient_function = lambda W : grad_cost_function(evaluate_y(giant_matrix, W*default_x), default_x, giant_matrix, W, reg_lambda)
 
-    optimized_result = sp.optimize.minimize(lambda_cost_function, start_W, method = "L-BFGS-B", jac=lambda_gradient_function)
+    #(lazy) derived optimization stuff for ln(W)
+    lnW_cost_function = lambda lnW : lambda_cost_function(np.exp(lnW))
+    lnW_gradient_function = lambda lnW : np.exp(lnW)*lambda_gradient_function(np.exp(lnW))
+
+    print("gradient at initial: ", lnW_gradient_function(lnW))
+    # optimization for W
+    # optimized_result = sp.optimize.minimize(lambda_cost_function, start_W, method = "L-BFGS-B", jac=lambda_gradient_function)
+
+    # optimization for lnW
+    optimized_result = sp.optimize.minimize(lnW_cost_function, lnW, method = "CG", jac=lnW_gradient_function)
     print(optimized_result)
-    return optimized_result['x']
+    return np.exp(optimized_result['x'])
     # x = default_x * W
     # for i in range(max_n_it):
     #     #compute y

### 2022-09-21 code/main.py

@@ -32,17 +32,22 @@ print(analysis.grad_cost_function(y_sim, default_x, pruned_giant_matrix, W, 1))
 
 
 #just some random value to try out gradient descent
-reg_lambda = 8.0
+reg_lambda = 1.0
+# reg_lambda = 0.0
 # reg_lambda = 1.0
 start_W = np.ones(default_x.shape)
-stepsize = 0.05
 
 print(np.linalg.svd(pruned_giant_matrix))
 
 
 # grad = analysis.grad_cost_function(y_sim, default_x, pruned_giant_matrix, W, reg_lambda)
 #
-# resulting_W = analysis.gradient_descent_algorithm(reg_lambda, np_sources, pruned_giant_matrix, start_W)
-# plt.figure()
-# plt.hist(resulting_W, bins=50)
-# plt.show()
+resulting_W = analysis.gradient_descent_algorithm(reg_lambda, np_sources, pruned_giant_matrix, start_W)
+y_corrected = analysis.evaluate_y(pruned_giant_matrix, default_x * resulting_W)
+
+plt.figure()
+plt.hist(np.log10(y_corrected), bins=50)
+plt.xlabel("log10(y_corrected)")
+plt.legend()
+plt.title("Simulated observations")
+plt.show()

### 2022-09-21 code/analysis.py
Optimization of the problem now seems to work, when changing to L-BFGS.

@@ -36,6 +36,32 @@ def grad_cost_function(y, x_default, giant_matrix, W, reg_lambda):
     reg_cost = reg_lambda / W * np.log(W)
     return (matrix_cost+reg_cost)
 
+#cost function using lnW
+def cost_function_ln(giant_matrix, xval_array, lnW, reg_lambda):
+    # W = np.exp(lnW)
+    lny = np.log(giant_matrix @ xval_array)
+    regularization = reg_lambda * lnW
+    print("min lnW: ", np.min(lnW))
+    print("max lnW: ", np.max(lnW))
+    cost = 1.0/2.0*(lny.T @ lny +regularization.T @ regularization)
+    print("cost: ", cost)
+    return cost
+
+
+#Gradient descent; is returned as a vector instead of a diagonal matrix
+#gradient function using lnW to derive to
+def grad_cost_function_ln(y, x_default, giant_matrix, lnW, reg_lambda):
+    # print("matrix shape_T: ", giant_matrix.T.shape)
+    # print("y shape: ", y.shape)
+    # print("matrix_T times y shape: ", (giant_matrix.T @ (np.log(y)/y)).shape)
+    # print("x_T shape: ", x_default.T.shape)
+    #to determine the diagonal of the large matrix, we do not explicitly need to compute this large matrix
+    # matrix_cost = np.diag((giant_matrix.T @ (np.log(y)/y)) @ x_default.T)
+    #this 'diagonal matrix' is stored in vector format
+    W = np.exp(lnW)
+    matrix_cost = W * (giant_matrix.T @ (np.log(y)/y)) * x_default
+    reg_cost = reg_lambda * lnW
+    return (matrix_cost+reg_cost)
 
 #returns the optimized weights
 def gradient_descent_algorithm(reg_lambda, np_sources, giant_matrix, start_W):
@@ -45,20 +71,26 @@ def gradient_descent_algorithm(reg_lambda, np_sources, giant_matrix, start_W):
     #optimizing lnW makes sure that W will always be positive
     lnW = np.log(W)
 
-    #original optimization stuff for W
-    lambda_cost_function = lambda W : cost_function(giant_matrix, W*default_x, W, reg_lambda)
-    lambda_gradient_function = lambda W : grad_cost_function(evaluate_y(giant_matrix, W*default_x), default_x, giant_matrix, W, reg_lambda)
+    # #original optimization stuff for W
[…]

### 2022-09-21 code/constants.py

@@ -2,7 +2,8 @@
 LIMIT_BACKWARD_TIME = 15
 N_SOURCES = 200
 N_DETECTIONS = 4636
-MAX_DATE = 40
+MAX_DATE = 365
+# MAX_DATE = 40
 #allowing us to use values for 1 entire year
 # N_COLS=365+LIMIT_BACKWARD_TIME
 # forty days might be sufficient

### 2022-09-21 code/main.py

@@ -37,7 +37,7 @@ reg_lambda = 1.0
 # reg_lambda = 1.0
 start_W = np.ones(default_x.shape)
 
-print(np.linalg.svd(pruned_giant_matrix))
+# print(np.linalg.svd(pruned_giant_matrix))
 
 
 # grad = analysis.grad_cost_function(y_sim, default_x, pruned_giant_matrix, W, reg_lambda)

### 2022-09-21 code/main.py
Added plots of the computed y with different penalization parameters.

@@ -11,6 +11,8 @@ giant_matrix = read.import_combined_matrix()
 print("done importing matrix")
 #and throw away all zero/negative rows
 pruned_giant_matrix = read.prune_matrix(giant_matrix)
+negative_rows_giant_matrix = read.negative_rows_of(giant_matrix)
+
 
 #default source assumes all sources to be constant at all times
 default_x = np.kron(np_sources, np.ones(cst.N_COLS))
@@ -49,5 +51,24 @@ plt.figure()
 plt.hist(np.log10(y_corrected), bins=50)
 plt.xlabel("log10(y_corrected)")
 plt.legend()
-plt.title("Simulated observations")
+plt.title("Simulated observations lambda: "+str(reg_lambda))
+# plt.show()
+
+
+#simulated non-detections after optimization
+y_nondetections = analysis.evaluate_y(negative_rows_giant_matrix, default_x * resulting_W)
+#also plotting a histogram of the non-detections
+plt.figure()
+plt.hist(np.log10(-y_nondetections), bins=50)
+plt.xlabel("log10(-y_nondetections)")
+plt.title("Simulated non-detections lambda: "+str(reg_lambda))
+
+plt.figure()
+plt.hist(np.log10(resulting_W), bins=50)
+plt.xlabel("log10(weights)")
+# plt.hist(resulting_W, bins=50)
+# plt.xlabel("weights")
+plt.title("Optimized weights lambda: "+str(reg_lambda))
+print("weights: ", resulting_W)
+
 plt.show()

### 2022-09-21 code/readin.py

@@ -138,6 +138,11 @@ def import_combined_matrix():
 def prune_matrix(matrix):
     return matrix[(matrix.min(axis=1)>=0.0) & (matrix.max(axis=1)>0.0)]
 
+
+#only keep negative row from the matrix
+def negative_rows_of(matrix):
+    return matrix[matrix.min(axis=1)<0.0]
+
 # for file in os.listdir(MATRIX_LOCATION):
 #     if file.endswith(".csv"):
 #         matrix_file = pandas.read_csv(os.path.join(MATRIX_LOCATION, file), header=None)

### 2022-09-21 geostd.py
maps

@@ -61,14 +61,11 @@ days3 = days2.iloc[range(start, stop + 1)].copy()
 samples3 = samples2[samples2['d'].isin(days3.index.values)].copy()
 #samples3.apply('_'.join, axis = 1).to_csv('48d8sta.csv')
 
+
+## get contributions for all 48 days
+total = []
 for day, day_stations in days3.iteritems():
     files = samples3[samples3['d'] == day].apply('_'.join, axis = 1)
-
-    ## sum influence from all stations
-    #for f in files:
-    #    data = pd.read_csv(os.path.join(m_paths, f), header = None)
-    #    # np.matmul(data.T, sources['scaled'].values)
-    #    z = data.sum(axis = 1) # total over 15 days
 
     ## sum influence from all stations
     daily = []
@@ -77,81 +74,83 @@ for day, day_stations in days3.iteritems():
         # np.matmul(data.T, sources['scaled'].values)
         dc = data[data > 0].sum(axis = 1) # total over 15 days, positive contributions only
         daily += [dc]
+    total += [daily]
 
-    z = np.array(daily).sum(axis = 0)
+# values
+z = np.array(total).sum(axis = 1).std(axis = 0) # mean, std, var
 
-    # color map
-    norm = mplc.Normalize(vmin = 0, vmax = np.percentile(z, 95))
-    cmap = cm.jet
-    m = cm.ScalarMappable(norm = norm, cmap = cmap)
-    colos = m.to_rgba(z)
-    
-    ## view
-    crs = ccrs.NorthPolarStereo()
-    crs_proj4 = crs.proj4_init
+# color map
+norm = mplc.Normalize(vmin = 0, vmax = np.percentile(z, 95))
+cmap = cm.jet
+m = cm.ScalarMappable(norm = norm, cmap = cmap)
+colos = m.to_rgba(z)
 
-    ## patches
-    points = MultiPoint([v for v in zip(sources['Longitude'], sources['Latitude'])])
-    regions = voronoi_diagram(points)
-    polygons = [v for v in regions.geoms]
-    gdf = gpd.GeoDataFrame(index = range(len(polygons)), geometry = polygons, crs = 'EPSG:4326') # EPSG 4326 - WGS 84
-    gdf2 = gdf.to_crs(crs_proj4) # project points to new CRS
+## view
[…]

### 2022-09-21 geostd.py
corrected patch colors
also fixed scale for colormaps

@@ -79,12 +79,6 @@ for day, day_stations in days3.iteritems():
 # values
 z = np.array(total).sum(axis = 1).std(axis = 0) # mean, std, var
 
-# color map
-norm = mplc.Normalize(vmin = 0, vmax = np.percentile(z, 95))
-cmap = cm.jet
-m = cm.ScalarMappable(norm = norm, cmap = cmap)
-colos = m.to_rgba(z)
-
 ## view
 crs = ccrs.NorthPolarStereo()
 crs_proj4 = crs.proj4_init
@@ -96,40 +90,26 @@ polygons = [v for v in regions.geoms]
 gdf = gpd.GeoDataFrame(index = range(len(polygons)), geometry = polygons, crs = 'EPSG:4326') # EPSG 4326 - WGS 84
 gdf2 = gdf.to_crs(crs_proj4) # project points to new CRS
 
-## notes on: index = range(len(polygons))
-# less polygons 196 than stations 200
-# POINT (49.47658 54.18786)     3
-# POINT (49.52358 54.20081)     3
-# 172   Russian Federation	VK-50	    54.187856	49.476581
-# 175   Russian Federation	SM-3	    54.187856	49.476581
-# 179   Russian Federation	MIR.M1	    54.187856	49.476581
-# 178   Russian Federation	BOR-60	    54.200812	49.523576
-# 196   Russian Federation	RBT-10/2    54.200812	49.523576
-# 197   Russian Federation	RBT-6	    54.200812	49.523576
-# => overlap
-## MANUAL FIX
-z = z.tolist()
-z[172] = z[172] + z[175] + z[179]
-z[178] = z[178] + z[196] + z[197]
-z = [v for k, v in enumerate(z) if k not in [175, 179, 178, 197]]
-colos = m.to_rgba(z)
-
-points = MultiPoint([v for k, v in enumerate(zip(sources['Longitude'], sources['Latitude'])) if k not in [175, 179, 178, 197]])
-regions = voronoi_diagram(points)
-polygons = [v for v in regions.geoms]
-gdf = gpd.GeoDataFrame(index = range(len(polygons)), geometry = polygons, crs = 'EPSG:4326') # EPSG 4326 - WGS 84
-gdf2 = gdf.to_crs(crs_proj4) # project points to new CRS
-# todo: group very small points or patches
-
 ## markers
 geometry = [Point(xy) for xy in zip(sources['Longitude'], sources['Latitude'])]
[…]

### 2022-09-21 read_geopandas.py

@@ -77,15 +77,8 @@ for day, day_stations in days3.iteritems():
         # np.matmul(data.T, sources['scaled'].values)
         dc = data[data > 0].sum(axis = 1) # total over 15 days, positive contributions only
         daily += [dc]
-
     z = np.array(daily).sum(axis = 0)
 
-    # color map
-    norm = mplc.Normalize(vmin = 0, vmax = np.percentile(z, 95))
-    cmap = cm.jet
-    m = cm.ScalarMappable(norm = norm, cmap = cmap)
-    colos = m.to_rgba(z)
-    
     ## view
     crs = ccrs.NorthPolarStereo()
     crs_proj4 = crs.proj4_init
@@ -96,42 +89,28 @@ for day, day_stations in days3.iteritems():
     polygons = [v for v in regions.geoms]
     gdf = gpd.GeoDataFrame(index = range(len(polygons)), geometry = polygons, crs = 'EPSG:4326') # EPSG 4326 - WGS 84
     gdf2 = gdf.to_crs(crs_proj4) # project points to new CRS
-
-    ## notes on: index = range(len(polygons))
-    # less polygons 196 than stations 200
-    # POINT (49.47658 54.18786)     3
-    # POINT (49.52358 54.20081)     3
-    # 172   Russian Federation	VK-50	    54.187856	49.476581
-    # 175   Russian Federation	SM-3	    54.187856	49.476581
-    # 179   Russian Federation	MIR.M1	    54.187856	49.476581
-    # 178   Russian Federation	BOR-60	    54.200812	49.523576
-    # 196   Russian Federation	RBT-10/2    54.200812	49.523576
-    # 197   Russian Federation	RBT-6	    54.200812	49.523576
-    # => overlap
-    ## MANUAL FIX
-    z = z.tolist()
-    z[172] = z[172] + z[175] + z[179]
-    z[178] = z[178] + z[196] + z[197]
-    z = [v for k, v in enumerate(z) if k not in [175, 179, 178, 197]]
-    colos = m.to_rgba(z)
-
-    points = MultiPoint([v for k, v in enumerate(zip(sources['Longitude'], sources['Latitude'])) if k not in [175, 179, 178, 197]])
-    regions = voronoi_diagram(points)
-    polygons = [v for v in regions.geoms]
-    gdf = gpd.GeoDataFrame(index = range(len(polygons)), geometry = polygons, crs = 'EPSG:4326') # EPSG 4326 - WGS 84
[…]

### 2022-09-21 code/analysis.py
Delete code directory

@@ -1,38 +1,0 @@
-import numpy as np
-
-def ln_y_squared(giant_matrix, xval_array):
-    lny = np.log(giant_matrix @ xval_array)
-    return lny.T @ lny
-
-def cost_function(giant_matrix, W, lambd):
-    # TODO: use W instead of xval_array; hint: reconstruct xvalues using this W (just kron as below)
-    lny = np.log(giant_matrix @ xval_array)
-    regularization = lambd * np.log(weigths)
-    return 1.0/2.0*(lny.T @ lny +regularization.T @ regularization)
-
-def evaluate_x(giant_matrix, xval_array):
-    return giant_matrix @ xval_array
-
-
-
-
-#METRICS
-# Caculate metrics
-# mean Model Bias
-def MB(results):
-    return np.mean(results[np.where(results>=0)])-1
-
-#Fractional Bias
-def FB(results):
-    return 2*((1-np.mean(results[np.where(results>=0)]))/(1+np.mean(results[np.where(results>=0)])))
-
-#Geometric Mean Bias
-def MG(results):
-    return np.exp(-np.mean(np.log(results[np.where(results>=0)])))
-
-# Fraction of the data where the model is within a fraction 2 correct
-def FAC2(results):
-    return np.size(np.where((results < 2) & (results > .5)))/np.size(np.where(results>=0)) # Fraction of the data where the model is within a
-
-#log10Error = np.mean(np.log10(results[np.where(results>=0)]))
-#log10Error = np.mean(np.log10(results[np.where(results>=0)]))

### 2022-09-21 code/constants.py

@@ -1,6 +1,0 @@
-#Just a file defining some constants for these scripts
-LIMIT_BACKWARD_TIME = 15
-N_SOURCES = 200
-N_DETECTIONS = 4636
-#allowing us to use values for 1 entire year
-N_COLS=365+LIMIT_BACKWARD_TIME

### 2022-09-21 code/main.py

@@ -1,23 +1,0 @@
-import readin as read
-import analysis as analysis
-import numpy as np
-import constants as cst
-import matplotlib.pyplot as plt
-
-np_sources = read.import_scaled_sources()
-print("importing matrix")
-#import the combined matrix
-giant_matrix = read.import_combined_matrix()
-print("done importing matrix")
-#and throw away all zero/negative rows
-pruned_giant_matrix = read.prune_matrix(giant_matrix)
-
-#default source assumes all sources to be constant at all times
-default_x = np.kron(np_sources, np.ones(cst.N_COLS))
-y_sim = analysis.evaluate_x(pruned_giant_matrix, default_x)
-
-plt.hist(np.log10(y_sim), bins=50)
-plt.xlabel("log10(y_sim)")
-plt.legend()
-plt.title("Simulated observations")
-plt.show()

### 2022-09-21 code/readin.py

@@ -1,243 +1,0 @@
-import pandas
-import os
-import numpy as np
-import scipy.optimize as opt
-import datetime
-import constants as cst
-
-MATRIX_LOCATION = "./../data/srs_csv_files/"
-# #TODO: figure out these parameters from the files themselves?
-# LIMIT_BACKWARD_TIME = 15
-# N_SOURCES = 200
-# N_DETECTIONS = 4636
-# #allowing us to use values for 1 entire year
-# N_COLS=365+LIMIT_BACKWARD_TIME
-
-
-# scalings=pandas.read_csv("./../data/scalings.csv", header=None)
-# sources=pandas.read_csv("./../data/data_xe133_NH_yearly.csv")#has a header file, not that we actually need it
-#
-# print(scalings)
-# print(sources)
-#
-# np_scalings=scalings[0].to_numpy()
-# np_sources=sources['emission'].to_numpy()*1000#emission in mBq instead of Bq
-# # print(np_sources.size)
-# # print(np_sources.shape)
-# #rescaling sources
-# # print("scalings: ", np_scalings)
-# # print("before dividing: ", np_sources)
-# np_sources=np_sources/np_scalings
-# # print("after dividing:", np_sources)
-# #np.divide(np_sources,np_scalings)
-#
-#
-# # print(np_scalings.size)
-# # print(np_scalings.shape)
-# # print("np sources: ", np_sources.size)
-#
-# n_datafiles=0
-# for file in os.listdir(MATRIX_LOCATION):
-#     if file.endswith(".csv"):
-#         n_datafiles+=1
-# results=np.zeros(n_datafiles)
-# print(results)
-
-
-
-
-#TODO figure out whether this thing needs to be sparse
-#probably
-# def fill_matrix(y, index, ):
-    # row =
-    # col =
-    # data =
-#or just construct from standard numpy array...
-
-def import_scalings():
-    scalings=pandas.read_csv("./../data/scalings.csv", header=None)
-    np_scalings=scalings[0].to_numpy()
-    return np_scalings
-
-def import_sources():
-    sources=pandas.read_csv("./../data/data_xe133_NH_yearly.csv")#has a header file, not that we actually need it
-    np_sources=sources['emission'].to_numpy()*1000#emission in mBq instead of Bq
-    return np_sources
-
-def import_scaled_sources():
-    return import_sources()/import_scalings()
-
-# index=0
-
-# #timesteps times number of emissions
-# nb_data_cols=np.size(np_sources)
-# print(nb_data_cols)
-
[…]

### 2022-09-21 geostd.py
multiplied by scaling factor
also higher res plots

@@ -73,7 +73,8 @@ for day, day_stations in days3.iteritems():
         data = pd.read_csv(os.path.join(m_paths, f), header = None)
         # np.matmul(data.T, sources['scaled'].values)
         dc = data[data > 0].sum(axis = 1) # total over 15 days, positive contributions only
-        daily += [dc]
+        sdc = np.multiply(dc, sources['scaled']) # scaled daily contributions
+        daily += [sdc]
     total += [daily]
 
 # values
@@ -106,7 +107,7 @@ poly_colors = [z[idx].sum() for idx in poly_pts]
 
 # color map
 #norm = mplc.Normalize(vmin = np.percentile(z, 5), vmax = np.percentile(z, 95))
-norm = mplc.Normalize(vmin = 0, vmax = 100)
+norm = mplc.Normalize(vmin = 0, vmax = 0.04)
 cmap = cm.jet
 m = cm.ScalarMappable(norm = norm, cmap = cmap)
 colos = m.to_rgba(z)

### 2022-09-21 read_geopandas.py

@@ -76,7 +76,8 @@ for day, day_stations in days3.iteritems():
         data = pd.read_csv(os.path.join(m_paths, f), header = None)
         # np.matmul(data.T, sources['scaled'].values)
         dc = data[data > 0].sum(axis = 1) # total over 15 days, positive contributions only
-        daily += [dc]
+        sdc = np.multiply(dc, sources['scaled']) # scaled daily contributions
+        daily += [sdc]
     z = np.array(daily).sum(axis = 0)
 
     ## view
@@ -106,13 +107,13 @@ for day, day_stations in days3.iteritems():
 
     # color map
     #norm = mplc.Normalize(vmin = np.percentile(z, 5), vmax = np.percentile(z, 95))
-    norm = mplc.Normalize(vmin = 0, vmax = 100)
+    norm = mplc.Normalize(vmin = 0, vmax = 0.04)
     cmap = cm.jet
     m = cm.ScalarMappable(norm = norm, cmap = cmap)
     colos = m.to_rgba(z)
     
     ## plot
-    fig, ax = plt.subplots(subplot_kw = {'projection': crs}, figsize = (16, 16))
+    fig, ax = plt.subplots(subplot_kw = {'projection': crs}, figsize = (40, 40))
     ax.add_geometries(world2['geometry'], crs = crs)
     gdf2.plot(ax = ax, marker = 'o', color = colos, alpha = 0.5, zorder = 10)
     gdf4.plot(ax = ax, marker = 'o', color = 'red', markersize = 5, zorder = 11)
@@ -141,5 +142,3 @@ for day, day_stations in days3.iteritems():
 # animation
 # https://underworldcode.github.io/stripy/2.0.5b2/FrontPage.html
 # https://coolum001.github.io/voronoi.html
-
-

### 2022-09-21 geostd.py
higher res plots

@@ -113,7 +113,7 @@ m = cm.ScalarMappable(norm = norm, cmap = cmap)
 colos = m.to_rgba(z)
 
 ## plot
-fig, ax = plt.subplots(subplot_kw = {'projection': crs}, figsize = (16, 16))
+fig, ax = plt.subplots(subplot_kw = {'projection': crs}, figsize = (40, 40))
 ax.add_geometries(world2['geometry'], crs = crs)
 gdf2.plot(ax = ax, marker = 'o', color = colos, alpha = 0.5, zorder = 10)
 gdf4.plot(ax = ax, marker = 'o', color = 'red', markersize = 5, zorder = 11)

### 2022-09-21 code/analysis.py
Implemented weighting for non-detections.

@@ -63,6 +63,61 @@ def grad_cost_function_ln(y, x_default, giant_matrix, lnW, reg_lambda):
     reg_cost = reg_lambda * lnW
     return (matrix_cost+reg_cost)
 
+
+#for all negative values, we transform it into the following
+def h_transform(Mx):
+    # print("before h_transform: ", Mx)
+    negative_indices = Mx < 0
+    # print("negative indices: ", negative_indices)
+    small_negative_indices = (Mx < 0) & (Mx > -2)
+    # print("small negative indices: ", small_negative_indices)
+    Mx[negative_indices] = -Mx[negative_indices]
+    Mx[small_negative_indices] = np.power(Mx[small_negative_indices], 2) / 4.0 + np.ones(Mx[small_negative_indices].shape)
+    # print("h_transform: ", Mx)
+    return Mx
+
+#returns gradient multiplication factors
+def grad_h_transform(Mx):
+    temp_grad = np.ones(Mx.shape)
+    negative_indices = Mx < 0
+
+    small_negative_indices = (Mx < 0) & (Mx > -2)
+    temp_grad[negative_indices] = -temp_grad[negative_indices]
+    temp_grad[small_negative_indices] = Mx[small_negative_indices] / 2.0
+    # print("tempgrad: ", temp_grad)
+    return temp_grad
+
+
+#cost function using lnW and the h function transform
+def cost_function_ln_hfun(giant_matrix, xval_array, lnW, reg_lambda):
+    # W = np.exp(lnW)
+    lny = np.log(h_transform(giant_matrix @ xval_array))
+    regularization = reg_lambda * lnW
+    print("min lnW: ", np.min(lnW))
+    print("max lnW: ", np.max(lnW))
+    cost = 1.0/2.0*(lny.T @ lny +regularization.T @ regularization)
+    print("cost: ", cost)
+    return cost
+
+
+#Gradient descent; is returned as a vector instead of a diagonal matrix
+#gradient function using lnW to derive to and using the h function transform
+#Note: y should be the original y=MWx, which contains negative values for the non-detections
+def grad_cost_function_ln_hfun(y, x_default, giant_matrix, lnW, reg_lambda):
+    # print("matrix shape_T: ", giant_matrix.T.shape)
+    # print("y shape: ", y.shape)
[…]

### 2022-09-21 code/main.py

@@ -12,6 +12,7 @@ print("done importing matrix")
 #and throw away all zero/negative rows
 pruned_giant_matrix = read.prune_matrix(giant_matrix)
 negative_rows_giant_matrix = read.negative_rows_of(giant_matrix)
+nonzero_giant_matrix = read.nonzero_rows_of(giant_matrix)
 
 
 #default source assumes all sources to be constant at all times
@@ -24,6 +25,16 @@ plt.legend()
 plt.title("Simulated observations")
 # plt.show()
 
+y_non = analysis.evaluate_y(negative_rows_giant_matrix, default_x)
+#also plotting the non-detections
+plt.figure()
+plt.hist(np.log10(-y_non), bins=50)
+plt.xlabel("log10(-y_non)")
+plt.legend()
+plt.title("Simulated non-detections")
+
+# plt.show()
+
 # nb_rows = giant_matrix.shape[1]
 print("giant matrix shape: ", pruned_giant_matrix.shape)
 # print("nb cols: ", nb_rows)
@@ -34,7 +45,7 @@ print(analysis.grad_cost_function(y_sim, default_x, pruned_giant_matrix, W, 1))
 
 
 #just some random value to try out gradient descent
-reg_lambda = 1.0
+reg_lambda = 0.1
 # reg_lambda = 0.0
 # reg_lambda = 1.0
 start_W = np.ones(default_x.shape)
@@ -44,8 +55,9 @@ start_W = np.ones(default_x.shape)
 
 # grad = analysis.grad_cost_function(y_sim, default_x, pruned_giant_matrix, W, reg_lambda)
 #
-resulting_W = analysis.gradient_descent_algorithm(reg_lambda, np_sources, pruned_giant_matrix, start_W)
-y_corrected = analysis.evaluate_y(pruned_giant_matrix, default_x * resulting_W)
+# resulting_W = analysis.gradient_descent_algorithm(reg_lambda, np_sources, pruned_giant_matrix, start_W)
+resulting_W = analysis.gradient_descent_algorithm(reg_lambda, np_sources, nonzero_giant_matrix, start_W)
+y_corrected = analysis.h_transform(analysis.evaluate_y(nonzero_giant_matrix, default_x * resulting_W))
 
 plt.figure()
 plt.hist(np.log10(y_corrected), bins=50)

### 2022-09-21 code/readin.py

@@ -143,6 +143,9 @@ def prune_matrix(matrix):
 def negative_rows_of(matrix):
     return matrix[matrix.min(axis=1)<0.0]
 
+def nonzero_rows_of(matrix):
+    return matrix[(matrix.max(axis=1)>0.0) | (matrix.min(axis=1)<0.0)]
+
 # for file in os.listdir(MATRIX_LOCATION):
 #     if file.endswith(".csv"):
 #         matrix_file = pandas.read_csv(os.path.join(MATRIX_LOCATION, file), header=None)

### 2022-09-21 code/constants.py
Fix incorrect number of files to read.

@@ -1,7 +1,7 @@
 #Just a file defining some constants for these scripts
 LIMIT_BACKWARD_TIME = 15
 N_SOURCES = 200
-N_DETECTIONS = 4636
+N_DETECTIONS = 4363
 MAX_DATE = 365
 # MAX_DATE = 40
 #allowing us to use values for 1 entire year

### 2022-09-21 code/main.py

@@ -12,7 +12,6 @@ print("done importing matrix")
 #and throw away all zero/negative rows
 pruned_giant_matrix = read.prune_matrix(giant_matrix)
 negative_rows_giant_matrix = read.negative_rows_of(giant_matrix)
-nonzero_giant_matrix = read.nonzero_rows_of(giant_matrix)
 
 
 #default source assumes all sources to be constant at all times
@@ -56,8 +55,8 @@ start_W = np.ones(default_x.shape)
 # grad = analysis.grad_cost_function(y_sim, default_x, pruned_giant_matrix, W, reg_lambda)
 #
 # resulting_W = analysis.gradient_descent_algorithm(reg_lambda, np_sources, pruned_giant_matrix, start_W)
-resulting_W = analysis.gradient_descent_algorithm(reg_lambda, np_sources, nonzero_giant_matrix, start_W)
-y_corrected = analysis.h_transform(analysis.evaluate_y(nonzero_giant_matrix, default_x * resulting_W))
+resulting_W = analysis.gradient_descent_algorithm(reg_lambda, np_sources, giant_matrix, start_W)
+y_corrected = analysis.h_transform(analysis.evaluate_y(giant_matrix, default_x * resulting_W))
 
 plt.figure()
 plt.hist(np.log10(y_corrected), bins=50)

### 2022-09-21 code/readin.py

@@ -143,8 +143,11 @@ def prune_matrix(matrix):
 def negative_rows_of(matrix):
     return matrix[matrix.min(axis=1)<0.0]
 
-def nonzero_rows_of(matrix):
-    return matrix[(matrix.max(axis=1)>0.0) | (matrix.min(axis=1)<0.0)]
+# def nonzero_rows_of(matrix):
+#     print("nb zero rows: ", matrix[(matrix.max(axis=1)==0.0) & (matrix.min(axis=1)==0.0)].shape)
+#     print("matrix shape: ", matrix.shape)
+#     print("returned matrix shape: ", matrix[(matrix.max(axis=1)>0.0) | (matrix.min(axis=1)<0.0)].shape)
+#     return matrix[(matrix.max(axis=1)>0.0) | (matrix.min(axis=1)<0.0)]
 
 # for file in os.listdir(MATRIX_LOCATION):
 #     if file.endswith(".csv"):

### 2022-09-22 code/analysis.py
Documented code a bit better

@@ -2,13 +2,16 @@ import numpy as np
 import constants as cst
 import scipy as sp
 
+#Note: a lot of unused functions are defined here.
+
+
+# cost function without regularization
 def ln_y_squared(giant_matrix, xval_array):
     lny = np.log(giant_matrix @ xval_array)
     return lny.T @ lny
 
-#W is vector
-#TODO FIXME
-#xval_array = W * default_x
+#cost function with regularization
+#xval_array = W * default_x, in which W is the vector form of the weights
 def cost_function(giant_matrix, xval_array, W, reg_lambda):
     # TODO: use W instead of xval_array; hint: reconstruct xvalues using this W (just kron as below)
     lny = np.log(giant_matrix @ xval_array)
@@ -17,18 +20,15 @@ def cost_function(giant_matrix, xval_array, W, reg_lambda):
     print("min weight: ", np.min(W))
     return 1.0/2.0*(lny.T @ lny +regularization.T @ regularization)
 
+#evaluate the result
+#xval_array = W * default_x, in which W is the vector form of the weights
 def evaluate_y(giant_matrix, xval_array):
     return giant_matrix @ xval_array
-
 
 
 #Gradient descent; is returned as a vector instead of a diagonal matrix
 #W is assumed to be a vector (corresponds with a diagonal matrix)
 def grad_cost_function(y, x_default, giant_matrix, W, reg_lambda):
-    # print("matrix shape_T: ", giant_matrix.T.shape)
-    # print("y shape: ", y.shape)
-    # print("matrix_T times y shape: ", (giant_matrix.T @ (np.log(y)/y)).shape)
-    # print("x_T shape: ", x_default.T.shape)
     #to determine the diagonal of the large matrix, we do not explicitly need to compute this large matrix
     # matrix_cost = np.diag((giant_matrix.T @ (np.log(y)/y)) @ x_default.T)
     #this 'diagonal matrix' is stored in vector format
@@ -51,10 +51,6 @@ def cost_function_ln(giant_matrix, xval_array, lnW, reg_lambda):
 #Gradient descent; is returned as a vector instead of a diagonal matrix
 #gradient function using lnW to derive to
 def grad_cost_function_ln(y, x_default, giant_matrix, lnW, reg_lambda):
-    # print("matrix shape_T: ", giant_matrix.T.shape)
-    # print("y shape: ", y.shape)
[…]

### 2022-09-22 code/constants.py

@@ -1,10 +1,19 @@
-#Just a file defining some constants for these scripts
+#This is a file defining some constants for the scripts in this folder
+#Note: technically these can just be determined by reading the csv files themselves
+
+#how many days the sensitivities go backwards in time
 LIMIT_BACKWARD_TIME = 15
+#number of sources
 N_SOURCES = 200
+#number of detections
 N_DETECTIONS = 4363
+#maximal date (in our case a single year)
 MAX_DATE = 365
+#if you want to run it using data from less days, just decrease the MAX_DATE
 # MAX_DATE = 40
-#allowing us to use values for 1 entire year
-# N_COLS=365+LIMIT_BACKWARD_TIME
-# forty days might be sufficient
+#total number of days to model (year + buffer for sensitivities going backwards in time)
 N_COLS=MAX_DATE+LIMIT_BACKWARD_TIME
+
+#constants you want to change
+#value for the regularization parameter; higher will for the weights closer to 1, while lower will allow more overfitting
+REG_LAMBDA=0.1

### 2022-09-22 code/main.py

@@ -4,82 +4,77 @@ import numpy as np
 import constants as cst
 import matplotlib.pyplot as plt
 
+#Note: in this file, basic analysis of the optimization method is done by looking at histograms.
+# More advanced methods/actual measures can also be implemented instead.
+
+
+#importing all data
 np_sources = read.import_scaled_sources()
 print("importing matrix")
 #import the combined matrix
 giant_matrix = read.import_combined_matrix()
 print("done importing matrix")
-#and throw away all zero/negative rows
-pruned_giant_matrix = read.prune_matrix(giant_matrix)
-negative_rows_giant_matrix = read.negative_rows_of(giant_matrix)
+#and seperate the matrix into the positive and negative rows respectively
+# this can be useful for plotting later on
+pruned_giant_matrix = read.prune_matrix(giant_matrix)#positive rows
+negative_rows_giant_matrix = read.negative_rows_of(giant_matrix)#negative rows
 
 
 #default source assumes all sources to be constant at all times
+#default_x = [N_COLS × source_0, N_COLS × source_1, …]^T
 default_x = np.kron(np_sources, np.ones(cst.N_COLS))
+#compute for comparison all simulated detections
 y_sim = analysis.evaluate_y(pruned_giant_matrix, default_x)
 
+#plot simulated detections
 plt.hist(np.log10(y_sim), bins=50)
 plt.xlabel("log10(y_sim)")
 plt.legend()
 plt.title("Simulated observations")
-# plt.show()
 
+#compute for comparison all simulated non-detections
 y_non = analysis.evaluate_y(negative_rows_giant_matrix, default_x)
-#also plotting the non-detections
+#also plot the non-detections
 plt.figure()
 plt.hist(np.log10(-y_non), bins=50)
 plt.xlabel("log10(-y_non)")
 plt.legend()
 plt.title("Simulated non-detections")
 
-# plt.show()
-
-# nb_rows = giant_matrix.shape[1]
-print("giant matrix shape: ", pruned_giant_matrix.shape)
-# print("nb cols: ", nb_rows)
-
-# print(analysis.grad_cost_function(np.ones(nb_rows)))
-W = np.ones(default_x.shape)
-print(analysis.grad_cost_function(y_sim, default_x, pruned_giant_matrix, W, 1))
-
-
-#just some random value to try out gradient descent
[…]

### 2022-09-22 notes/notes.tex
incorporate h-transformation into notes (messy)

@@
     \item $\odot$ is the Hadamard (i.e.\, elementwise) product, given by $(AB)_{ij} = A_{ij} B_{ij}$.
+    \item $\mathrm{vec}: \mathbb{R}^{p \times q} \rightarrow \mathbb{R}^{pq}$ is the vectorisation operator. It can be understood as stacking the columns of a matrix in one long vector. 
     \item $\mathrm{diag}$ is the canonical identification of $p$-vectors and $p \times p$ diagonal matrices, i.e.,
@@
         A standard rule of thumb in numerical linear algebra is that the Kronecker product serves a mostly theoretical purpose and is rarely calculated explicitly.
-See, e.g.
-\cite[Section 12.3]{Golub2013}.
+        This is because the Kronecker product has the following convenient property \cite[Section 12.3]{Golub2013}: $\mathrm{vec} (AXB) = (B^T \otimes A) \mathrm{vec} X$ for all matrices $A,X,B$.
+    \item For a vector-valued function $F(x)$, the Jacobian matrix is denoted as $\frac{\partial F}{\partial x}$.
 \end{itemize}
@@
 
-\subsection{Fitting the model}
+\section{Fitting the model}
+\subsection{Regularisation}
 The indices $i = 1,\dots,n$ are partitioned as a disjoint union $\mathcal{I}_{det} \cup \mathcal{I}_{0}$ where $\mathcal{I}_0$ contains the indices of non-detection events.
@@
 
+\subsection{Constrained optimisation (similar idea as regularisation)}
+Constrained optimisation is similar to using regularisation.
+Suppose that the number of components in $\tilde{x}$ is $m T$ (where $m$ is the number of sources and $T$ is the number of days).
+If we want to bound the average of $(\log w_{ij})^2$ by some value $\delta^2$, then we have the constrained optimisation problem
+\begin{align*}
+    \min\quad & \frac{1}{2} \norm{\log.(P_{det} M_{shift} W \tilde{x})}^2 \\
+    \text{s.t.}\quad & \frac{1}{mT} \norm{\log W}^2 \le \delta^2
+\end{align*} 
+or equivalently 
+\begin{align*}
+    \min\quad & \frac{1}{2} \norm{\log.(P_{det} M_{shift} e^{\mathrm{diag}\, v} \tilde{x})}^2 \\
+    \text{s.t.}\quad & \frac{1}{mT} \norm{v}^2 \le \delta^2
+\end{align*}
+which can be solved using very similar ideas as the regularisation section.
+
+\section{Correcting for non-detections}
+
+To correct for non-detections, we apply the transformation $h$ from \cite{deMeutter2022uncertainty}. This is given by the continuously differentiable function
+$$
+h(y) = \begin{cases}
+    \frac{y^2}{4MDC} + MDC &\quad \text{if } y \le 2MDC \\
+    y &\quad \text{else}
+\end{cases}
+.$$
+This transformation maps the interval $[0, MDC]$ onto $[MDC, 2MDC]$.
+For observations that are significantly larger than $MDC$, this is approximately the identity.
+If $y$ and the simulated value $\hat{y}$ are both larger than $2MDC$, then $h(\hat{y}) / h(y) = \hat{y} / y$.
+If $y = 0$ because the concentration is not detectable, then we want a simulated value $\hat{y}$ such that $h(\hat{y}) \approx MDC$.
+In both cases, we can use the value of $h(\hat{y}) / h(y)$ as an indication of the accuracy of the simulation.
+
+In the dataset that we are working with, we know the values of $- \frac{M_{i,(j,\Delta)}}{MDC}$ for every non-detection event.
+In these events, we work as if $MDC = 1$.
+If the predicted value is $\hat{y}_i < 0$, we have
+$$
+h\left(\hat{y}\right) = \begin{cases}
+    \frac{\hat{y}^2}{4} + 1 &\quad\text{if } \hat{y} \le -2 \\
+    -\hat{y} &\quad\text{else}
+\end{cases}
+$$
+and $h(y) = 1$. For the detection events, we do not have the $MDC$ and cannot calculate $h$ exactly. For the detection events, we can approximate $h$ by the identity map.
+
+Now a good cost function is
+$$
+\min_{v} f(v) =
+\underbrace{\frac{1}{2}\norm{\log h.(\hat{y})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{v}^2}_{\text{deviation from average emission}}
+\quad\text{where}\quad 
+\hat{y} = M_{shift} e^{\mathrm{diag}\, v} \tilde{x}
+$$
+where $\log h.$ is the elementwise application of $x \mapsto \log h(x)$ to a vector.
+
+\begin{proposition}
+    Let $g$ be the vector defined elementwise by 
+    $$
+    g_i = \frac{\log h(\hat{y}_i)}{h(\hat{y}_i)} \frac{\mathrm{d}}{\mathrm{d}\hat{y}_i} h(\hat{y}_i)
+    .$$
+    Then 
+    $$
+    \mathrm{grad}\, f(v) = e^v \odot (M_{shift}^T g) \odot \tilde{x} 
+    $$
+\end{proposition}
+\begin{proof}
+
+Let $U = \mathrm{vec} \circ \mathrm{diag}$. Then $\mathrm{vec}\, W(v) = U e^v$ where $e^v$ denotes elementwise exponentiation.
+Define $$
+F(v) := \log h.(\hat{y}(v))
+.$$
+Since 
+$$
+\hat{y}(v) = M_{shift} W(v) \tilde{x}
+=
+(\tilde{x}^T \otimes M_{shift}) \mathrm{vec} W(v)
+=
+(\tilde{x}^T \otimes M_{shift}) U e^v
+,$$
+we have
+$$
+\frac{\partial F}{\partial v}
+=
+\mathrm{diag}(d \log h.(\hat{y}(v))) \frac{\partial \hat{y}}{\partial v}
+=
+\underbrace{\mathrm{diag}(d \log h.(\hat{y}(v)))}_H U^T (\tilde{x} \otimes M_{shift}^T) e^{\mathrm{diag}\, v}
+.$$
+Using standard differentiation rules for the squared norm of a vector-valued function, we obtain
+\begin{align*}
+\mathrm{grad}_v\, \left(\frac{1}{2} \norm{F(v)}^2 \right) =
+\left(\frac{\partial F}{\partial v}\right)^T F(v)
+&= e^{\mathrm{diag}\, v} U^T (\tilde{x} \otimes M_{shift}^T) H \log h.(\hat{y})
+\end{align*}
+Since $U$ is the unitary map that maps a vector onto (the vectorisation of) the corresponding diagonal matrix, $U^T$ maps the vectorisation of a matrix onto its diagonal.
+If $\mathrm{Diag}: \mathbb{R}^{p \times p} \rightarrow \mathbb{p}$ maps a matrix to its diagonal, we have
+\begin{align*}
+\mathrm{grad}_v\, \left(\frac{1}{2} \norm{F(v)}^2 \right)
+&=
+e^{\mathrm{diag}\, v} \mathrm{Diag}(M_{shift}^T H \log h.(\hat{y}) \tilde{x}^T) \\
+&=
+e^{\mathrm{diag}\, v} \left((M_{shift}^T H \log h.(\hat{y})) \odot \tilde{x}\right) \\
+&=
+e^{v} \odot \left(M_{shift}^T H \log h.(\hat{y})\right) \odot \tilde{x} \\
+&=
+e^v \odot (M_{shift}^T g(\hat{y})) \odot \tilde{x}
+\end{align*}
+where $g$ is the elementwise application of $x \mapsto \log h(x) \frac{\mathrm{d}}{\mathrm{d}x}(\log h(x))$. Thus, 
+$$
+g(\hat{y}_i) = \frac{\log h(\hat{y}_i)}{h(\hat{y}_i)} \frac{\mathrm{d}}{\mathrm{d}\hat{y}_i} h(\hat{y}_i)
+$$
+
+The foregoing gives the derivative of the first term in $f(W)$. The gradient of the second term is simply $\lambda v$.
+
+\end{proof}
+
+\section{Further work}
+\begin{itemize}
+    \item For certain emission sources in certain periods, there is an expected correlation between the emission at time $t$ and at time $t + 1, t + 2, \dots$.
+We can add a constraint on the autocorrelation.
+    \item Change the parametrisation to something like this: 
+    \begin{align*}
+        x_1^{2 \text{ Jan } 2014} &= \alpha_1^{(1)} x_1^{1 \text{ Jan } 2014} \\
+        x_1^{3 \text{ Jan} 2014} &= \alpha_2^{(1)} x_1^{2 \text{ Jan } 2014} \\
+        &\vdots
+    \end{align*}
+    with a constraint like $\alpha_{\min} \le \prod_{t=1}^T \alpha_t^{(j)} \le \alpha_{\max}$ for all $T$ and all $j = 1,\dots,m$.
+However, this would be difficult to implement.
+In addition, the complexity of inequality constrained optimisation can be combinatorial in the number of constraints if an active set method is used \cite{Nocedal2006}.
+
+    \item How do you store the sparse matrix $M_{shift}$ in a computer? Most programming languages have data structures for large sparse matrices?
+\end{itemize}
+
 \printbibliography

### 2022-09-22 code/main.py
Now also computes metrics for the optimized results.

@@ -54,7 +54,6 @@ plt.hist(np.log10(y_corrected), bins=50)
 plt.xlabel("log10(y_corrected)")
 plt.legend()
 plt.title("Simulated observations lambda: "+str(cst.REG_LAMBDA))
-# plt.show()
 
 
 #compute the simulated non-detections after optimization
@@ -76,5 +75,18 @@ plt.title("Optimized weights lambda: "+str(cst.REG_LAMBDA))
 np.savetxt("x_optimized_lambda_"+str(cst.REG_LAMBDA)+".csv", default_x * resulting_W)
 np.savetxt("W_optimized_lambda_"+str(cst.REG_LAMBDA)+".csv", resulting_W)
 
+# metrics for the detections with optimized weights
+print("Mean Bias: ", analysis.MB(y_corrected))
+print("Fractional Bias: ", analysis.FB(y_corrected))
+print("Geometric Mean Bias: ", analysis.MG(y_corrected))
+print("Data within factor 2: ", analysis.FAC2(y_corrected))
+#also print metrics to file
+printstring = ""
+printstring += "Mean Bias: "+ str(analysis.MB(y_corrected)) + "\n"
+printstring += "Fractional Bias: "+ str(analysis.FB(y_corrected)) + "\n"
+printstring += "Geometric Mean Bias: "+ str(analysis.MG(y_corrected)) +"\n"
+printstring += "Data within factor 2: "+ str(analysis.FAC2(y_corrected)) +"\n"
+with open("metrics lambda="+str(cst.REG_LAMBDA)+".txt", "w") as writefile:
+    writefile.write(printstring)
 
 plt.show()

### 2022-09-22 notes/notes.tex
restructure and clarify notes

@@
 In this expression, the unknown is $W$ and should be fit to the data.
-% There are a plethora of numerical techniques for solving linear equations where the unknown is a matrix.
-% In the following, we provide a theoretical description of a system that can be solved for $W$.
 
-% We can put constraints on $W$ of the form $W \in \mathcal{W}$, where $\mathcal{W}$ is some set of constrained diagonal matrices.
-Suppose we have a parametrisation $\Gamma: \mathbb{R}^{k} \rightarrow \mathcal{W}, u \mapsto \Gamma(u)$.
-% In this case, \cref{eq: time-evolving model concise} can be reformulated as a linear operation applied to $\Gamma(u)$ as follows:
-% $$
-% \hat{y} = ((S^{-1} x \otimes \vb{1}_{365})^T \otimes M_{shift}) \mathrm{vec}\, \Gamma(u)
-% .$$
-% If $\Gamma$ is a linear map, the above is a linear expression in $u$. Then, the minimisation of $\norm{y - \hat{y}}^2$ can be solved using numerical linear algebra.
-
-\section{Fitting the model}
-\subsection{Regularisation}
-The indices $i = 1,\dots,n$ are partitioned as a disjoint union $\mathcal{I}_{det} \cup \mathcal{I}_{0}$ where $\mathcal{I}_0$ contains the indices of non-detection events.
-Let $P_{det}$ be the linear map that removes the rows of a matrix that correspond to $\mathcal{I}_0$.
-We can measure the accuracy as $\norm{\log.(P_{det} \hat{y})}^2$ where $\log.(\cdot)$ is the elementwise \emph{natural} logarithm.
-The simplicity of $W$ can be measured as $\norm{\log W}^2$.
-If we forget about non-detections, we have the cost function
-$$
-\min_{W \text{ diagonal}} f(W) =
-\underbrace{\frac{1}{2}\norm{\log.(P_{det} M_{shift} W \tilde{x})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{\log W}^2}_{\text{deviation from average emission}}
-.$$
-In order to optimise $f$ numerically, we need to be able to measure changes to $W$.
-We do this implicitly by parametrising $W = e^{\mathrm{diag}\, v}$ for some vector $v$.
-If we apply an absolute change $v' := v + \Delta v$ in numerical optimisation, this corresponds to the relative change $W' = W e^{\mathrm{diag} \Delta v}$.
-Reformulating the cost function gives 
-$$
-\min_{v} f(v) =
-\underbrace{\frac{1}{2}\norm{\log.(P_{det} M_{shift} e^{\mathrm{diag}\, v} \tilde{x})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{v}^2}_{\text{deviation from average emission}}
-.$$
-
-\begin{proposition}
-    Let $\hat{z}$ be the vector $[\hat{z}_1\,\,\dots\,\,\hat{z}_n]^T$ where 
-    $$
-    \hat{z}_i = \begin{cases}
-        \frac{\log \hat{y}_i}{\hat{y}_i} & \text{if observation $i$ is a detection} \\
-        0 & \text{if observation $i$ is a non-detection} 
-    \end{cases}
-    .$$
-    Then
-    $$
-    \mathrm{grad}_{v} f(v) = e^v \odot (M_{shift}^T \hat{z}) \odot \tilde{x} + \lambda v
-    .$$
-\end{proposition}
-\begin{proof}
-
-Let $U = \mathrm{vec} \circ \mathrm{diag}$. Then $\mathrm{vec}\, W(v) = U e^v$ where $e^v$ denotes elementwise exponentiation.
-Define $$
-G(v) :=
-P_{det} \hat{y}_i
-=
-P_{det} M_{shift} W(v) \tilde{x}
-\quad\text{and}\quad
-F(v) := \log.(G(v))
-.$$
-Then 
-$$
-G(v) = P_{det} M_{shift} W(v) \tilde{x}
-=
-(\tilde{x}^T \otimes P_{det} M_{shift}) \mathrm{vec} W(v)
-=
-\underbrace{(\tilde{x}^T \otimes P_{det}M_{shift}) U}_A e^v
-$$
-so that 
-$$
-\frac{\partial G}{\partial v} = A e^{\mathrm{diag}\, v}
-.$$
-Differentiating $F(v) = \log.(G(v))$ gives:
-$$
-\frac{\partial F}{\partial v}
-=
-\frac{\partial}{\partial v} \log.(G(v))
-=
-\mathrm{diag}\,(G(v))^{-1} \frac{\partial G}{\partial v}
-=
-\mathrm{diag}\,(G(v))^{-1} A e^{\mathrm{diag}\, v }
-.$$
-Write $z := P_{det} \hat{z}$.
-We can interpret $z$ as the vector obtained by applying the map $x \mapsto (\log x) / x$ elementwise to $G(v)$.
-Using standard differentiation rules for the squared norm of a vector-valued function, we obtain
-\begin{align*}
-\mathrm{grad}_v\, \left(\frac{1}{2} \norm{F(v)}^2 \right) =\left(\frac{\partial F}{\partial v}\right)^T F(v)
-&= e^{\mathrm{diag}\, v} A^T \mathrm{diag}(G(t))^{-1} \log.(G(t)) \\
-&= e^{\mathrm{diag}\, v} A^T z \\
-&= e^{\mathrm{diag}\, v} U^T (\tilde{x} \otimes (P_{det} M_{shift})^T) z \\
-&= e^{\mathrm{diag}\, v} U^T (P_{det} M_{shift})^T z \tilde{x}^T \\
-&= e^{\mathrm{diag}\, v} U^T M_{shift}^T P_{det}^T P_{det} \hat{z} \tilde{x}^T \\
-&= e^{\mathrm{diag}\, v} U^T M_{shift}^T \hat{z} \tilde{x}^T
-\end{align*}
-where the last equality follows from the fact that $P_{det}^T P_{det}$ is the map that sets the rows corresponding to the non-detection events to zero. 
-By construction of $U$, the map $U^T$ projects its argument orthogonally onto the diagonal matrices and gives the diagonal element.
-The orthogonal projection of a matrix onto the diagonal matrices is obtained by setting the off-diagonal elements to zero.
-The diagonal elements of $M_{shift}^T \hat{z} \tilde{x}^T$ are $(M_{shift}^T \hat{z}) \odot \tilde{x}$, so that the above simplifies to
-$$
-\mathrm{grad}_v\, \left(\frac{1}{2} \norm{F(v)}^2 \right)
-=
-e^{\mathrm{diag}\, v} \left((M_{shift}^T \hat{z}) \odot \tilde{x}\right)
-=
-e^v \odot (M_{shift}^T \hat{z}) \odot \tilde{x}
-.$$
-
-The foregoing gives the derivative of the first term in $f(W)$. The gradient of the second term can is simply $\lambda v$.
-
-\end{proof}
-For even more evidence, there is a Julia script in the repository that checks this formula against the result obtained from automatic differentiation (using simplified mock data).
-The result is correct up to machine precision.
-
-\subsection{Constrained optimisation (similar idea as regularisation)}
-Constrained optimisation is similar to using regularisation.
-Suppose that the number of components in $\tilde{x}$ is $m T$ (where $m$ is the number of sources and $T$ is the number of days).
-If we want to bound the average of $(\log w_{ij})^2$ by some value $\delta^2$, then we have the constrained optimisation problem
-\begin{align*}
-    \min\quad & \frac{1}{2} \norm{\log.(P_{det} M_{shift} W \tilde{x})}^2 \\
-    \text{s.t.}\quad & \frac{1}{mT} \norm{\log W}^2 \le \delta^2
-\end{align*} 
-or equivalently 
-\begin{align*}
-    \min\quad & \frac{1}{2} \norm{\log.(P_{det} M_{shift} e^{\mathrm{diag}\, v} \tilde{x})}^2 \\
-    \text{s.t.}\quad & \frac{1}{mT} \norm{v}^2 \le \delta^2
-\end{align*}
-which can be solved using very similar ideas as the regularisation section.
-
-\section{Correcting for non-detections}
+\section{Correction for non-detections}
 
@@
 
-Now a good cost function is
+\section{Fitting the model}
+In the above, we argued that, the prediction is accurate if $h(\hat{y})$ is close to one, where $\hat{y}$ is the predicted concentration.
+This is equivalent to $\log^2(h(\hat{y}))$ being small.
+Notice that this penalises overestimates by a factor $\alpha$ as much as underestimates by a factor $\alpha$.
+That is,
+$$
+\log^2(\alpha) = \log^2(\alpha^{-1})
+.$$
+The average prediction error over all predictions is $\sum_{i=1}^n \log^2(h(\hat{y}_i)) = \norm{\log h(\hat{y})}^2$. This will be the first criterion for optimality of the model.
+
+The second criterion is the ratio between the daily emissions and the average emission for each source, i.e.
+the factors $W$.
+In a similar vein as the above, the size of $W$ can be measured as $\norm{\log W}^2$ where $\norm{\cdot}$ is the Frobenius norm\footnote{Since $W$ is a diagonal matrix, the matrix logarithm $\log W$ is simply the diagonal matrix with $\log w_{1,1}, \log w_{1,2}, \dots$ on the diagonal.
+Hence, $\norm{\log W}^2 = \log^2 w_{1,1} + \log^2 w_{1,2} + \dots$}.
+
+To balance these two criteria, we measure the quality of the model as 
+\begin{equation}
+    \label{eq: cost function regularised}
+f(W) := 
+\underbrace{\frac{1}{2}\norm{\log h.(\hat{y})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{\log W}^2}_{\text{deviation from average emission}}
+\quad\text{where}\quad 
+\hat{y} = M_{shift} W \tilde{x}
+\end{equation}
+where $\lambda \ge 0$ is a regularisation parameter.
+
+To optimise $f$ numerically, we will parametrise $W$ as $W = e^{\mathrm{diag}\, v}$ where $v = [v_1, v_2,\dots]$ is some vector and $e^{\mathrm{diag}\, v}$ is the diagonal matrix with $[e^{v_1}, e^{v_2},\dots]$ on the diagonal.
+The motivation for doing this is twofold:
+\begin{enumerate}
+    \item It ensures that $W$ remains strictly positive. In particular, the diagonal elements of $W$ can only tend towards zero if $v_j \rightarrow -\infty$ for some $i$. 
+    \item It implicitly measures changes to $W$ in a relative sense.
+If the optimisation algorithm performs a linear update $v_{k+1} = v_k + \Delta v$, this corresponds to a relative update $W_{k+1} = W_k e^{\mathrm{diag}\, \Delta v}$.
+\end{enumerate}
+
+Thus, in practice, we are interested in the optimisation problem
 $$
 \min_{v} f(v) =
-\underbrace{\frac{1}{2}\norm{\log h.(\hat{y})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{v}^2}_{\text{deviation from average emission}}
+\frac{1}{2}\norm{\log h.(\hat{y})}^2 + \frac{1}{2}\lambda \norm{v}^2
 \quad\text{where}\quad 
 \hat{y} = M_{shift} e^{\mathrm{diag}\, v} \tilde{x}
-$$
-where $\log h.$ is the elementwise application of $x \mapsto \log h(x)$ to a vector.
-
+.$$
+Most optimisation algorithms require at least the gradient of $f$. This is given by the following:
 \begin{proposition}
-    Let $g$ be the vector defined elementwise by 
+    Let $g = [g_1, \dots, g_n]$ be the vector defined elementwise by 
     $$
@@
     $$
+    where $e^v$ denotes elementwise exponentiation and $\odot$ is the Hadamard (elementwise) product.
 \end{proposition}
@@
 We can add a constraint on the autocorrelation.
-    \item Change the parametrisation to something like this: 
+    \item Time-correlation could be measured by parameterising the model as follows: 
     \begin{align*}
@@
 
-    \item How do you store the sparse matrix $M_{shift}$ in a computer? Most programming languages have data structures for large sparse matrices?
+    \item For the sake of simplicity, the current implementation does not store $M_{shift}$ in memory in a data structure for large and sparse matrices.
+This could be changed to improve memory use and/or computation time.
+
+    \item Other numerical optimisation algorithms may be more efficient or have better convergence properties.
+For instance, the Gauss-Newton and Levenberg-Marquardt methods are specifically designed for optimising the squared norm of a vector-valued function, but may require a large amount of memory \cite[Section 10.3]{Nocedal2006}.
+
+    \item Standard software libraries support optimisation with constraints. This is an alternative to regularisation and may be more interpretable.
+    Suppose that the number of components in $\tilde{x}$ is $m T$ (where $m$ is the number of sources and $T$ is the number of days).
+    If we want to bound the average of $(\log w_{ij})^2$ by some value $\delta^2$, then we have the constrained optimisation problem
+    \begin{align*}
+        \min\quad & \frac{1}{2} \norm{\log h.(\hat{y})}^2 \\
+        \text{s.t.}\quad & \frac{1}{mT} \norm{\log W}^2 \le \delta^2
+    \end{align*} 
+    which can be solved using similar ideas as those developed above.
+The prototypical constrained optimisation method (such as an active set method) would iterate towards a critical point of the Lagrangian
+    $$
+    \mathcal{L}(W, \lambda) = \frac{1}{2} \norm{\log h.(\hat{y})}^2 + \lambda\left(\frac{1}{mT} \norm{\log W}^2 - \delta^2\right)
+    $$
+    in which case the computational cost would be comparable to optimising \cref{eq: cost function regularised}.
+A more detailed explanation of constrained optimisation and the Lagrangian is found in \cite{Nocedal2006}.
 \end{itemize}

### 2022-09-22 notes/notes.tex
add discussion about optimality

@@
 f(W) := 
-\underbrace{\frac{1}{2}\norm{\log h.(\hat{y})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{\log W}^2}_{\text{deviation from average emission}}
+\underbrace{\frac{1}{2}\norm{\log h.(\hat{y}(W))}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{\log W}^2}_{\text{deviation from average emission}}
 \quad\text{where}\quad 
-\hat{y} = M_{shift} W \tilde{x}
+\hat{y}(W) = M_{shift} W \tilde{x}
 \end{equation}
@@
 
-To optimise $f$ numerically, we will parametrise $W$ as $W = e^{\mathrm{diag}\, v}$ where $v = [v_1, v_2,\dots]$ is some vector and $e^{\mathrm{diag}\, v}$ is the diagonal matrix with $[e^{v_1}, e^{v_2},\dots]$ on the diagonal.
+Observe that the model where $W$ is the identity matrix $\mathds{I}$ is equivalent to the original model. Furthermore, the regularisation term vanishes if and only if $W = \mathds{I}$. That is,
+$$
+f(\mathds{I}) = \frac{1}{2}\norm{\log h.(\hat{y}(\mathds{I}))}^2
+$$
+is the prediction error for the original model.
+For $W \ne \mathds{I}$, the regularisation term is strictly positive. This means that, if $f(W) \le f(\mathds{I})$, then the prediction error is decreased, i.e.,
+$$
+\frac{1}{2}\norm{\log h.(\hat{y}(W))}^2
+\le
+\frac{1}{2}\norm{\log h.(\hat{y}(\mathds{I}))}^2
+.$$
+Hence, the model corresponding to the minimum of \cref{eq: cost function regularised} predicts the data at least as accurately as the original model.
+
+To optimise $f$ numerically, we will parametrise $W$ in \cref{eq: cost function regularised} as $W = e^{\mathrm{diag}\, v}$ where $v = [v_1, v_2,\dots]$ is some vector and $e^{\mathrm{diag}\, v}$ is the diagonal matrix with $[e^{v_1}, e^{v_2},\dots]$ on the diagonal.
 The motivation for doing this is twofold:
@@
 .$$
+
 Most optimisation algorithms require at least the gradient of $f$. This is given by the following:

### 2022-09-22 notes/notes.tex
clarify proof of gradient

@@
     \item Measured concentrations $y_1,\dots,y_n$ where $n = 4636$. These are consecutive measurements for several measurement stations. 
-    \item $M_{i,(j,\Delta)}$ is a block in the sensitivity matrix $M$ that represents the contribution of $x_j$ from $\Delta - 1$ days ago to the observed concentration $y_i$.
+    \item $M_{i,(j,\Delta)}$ is a block in the sensitivity matrix $M$ that represents the contribution of $x_j$ from $\Delta$ days ago to the observed concentration $y_i$.
     \item Scaling factors $s_1,\dots,s_m$ (in \texttt{scalings.csv}). These are put in a diagonal matrix $S$.
@@
         A standard rule of thumb in numerical linear algebra is that the Kronecker product serves a mostly theoretical purpose and is rarely calculated explicitly.
-        This is because the Kronecker product has the following convenient property \cite[Section 12.3]{Golub2013}: $\mathrm{vec} (AXB) = (B^T \otimes A) \mathrm{vec} X$ for all matrices $A,X,B$.
+        This is because the Kronecker product satisfies $\mathrm{vec} (AXB) = (B^T \otimes A) \mathrm{vec} X$ for all matrices $A,X,B$ \cite[eq. 1.3.6]{Golub2013}.
     \item For a vector-valued function $F(x)$, the Jacobian matrix is denoted as $\frac{\partial F}{\partial x}$.
@@
 Thus, in practice, we are interested in the optimisation problem
-$$
-\min_{v} f(v) =
+\begin{equation}
+    \label{eq: cost function parameterised}
+\min_{v} f(v)
+\quad\text{where}\quad 
+f(v)
+=
 \frac{1}{2}\norm{\log h.(\hat{y})}^2 + \frac{1}{2}\lambda \norm{v}^2
-\quad\text{where}\quad 
+\quad\text{and}\quad 
 \hat{y} = M_{shift} e^{\mathrm{diag}\, v} \tilde{x}
-.$$
+.\end{equation}
 
@@
 \begin{proof}
-
-Let $U = \mathrm{vec} \circ \mathrm{diag}$. Then $\mathrm{vec}\, W(v) = U e^v$ where $e^v$ denotes elementwise exponentiation.
+Let $W = e^{\mathrm{diag}\, v}$ and let $U$ be the matrix corresponding to the linear map $\mathrm{vec} \circ \mathrm{diag}$. Then $\mathrm{vec}\, W(v) = U e^v$.
 Define $$
@@
 .$$
-Since 
+By using basic properties of the Kronecker product \cite[eq. 1.3.6]{Golub2013}, we have
 $$
@@
 ,$$
-we have
+so that
 $$
@@
 =
-\mathrm{diag}(d \log h.(\hat{y}(v))) \frac{\partial \hat{y}}{\partial v}
+\underbrace{\mathrm{diag}\left(
+    \left.\frac{\mathrm{d}}{\mathrm{d}x} \log h(x)\right\vert_{x = \hat{y}(v)}
+    \right)}_{:= H}
+\frac{\partial \hat{y}}{\partial v}
 =
-\underbrace{\mathrm{diag}(d \log h.(\hat{y}(v)))}_H U^T (\tilde{x} \otimes M_{shift}^T) e^{\mathrm{diag}\, v}
+H (\tilde{x}^T \otimes M_{shift}) U e^{\mathrm{diag}\, v}
+.$$
+The matrix $H$ appearing in this expression is defined by the diagonal elements 
+$$
+H_{ii} = \frac{1}{h(\hat{y}_i)}\left(\left.\frac{\mathrm{d}}{\mathrm{d}x} h(x)\right\vert_{x = \hat{y}_i} \right)
 .$$
@@
 \begin{align*}
-\mathrm{grad}_v\, \left(\frac{1}{2} \norm{F(v)}^2 \right) =
-\left(\frac{\partial F}{\partial v}\right)^T F(v)
-&= e^{\mathrm{diag}\, v} U^T (\tilde{x} \otimes M_{shift}^T) H \log h.(\hat{y})
+\mathrm{grad}_v\, \left(\frac{1}{2} \norm{F(v)}^2 \right)
+&=
+\left(\frac{\partial F}{\partial v}\right)^T F(v) \\
+&= e^{\mathrm{diag}\, v} U^T (\tilde{x} \otimes M_{shift}^T) H \log h.(\hat{y}) \\
+&= e^{\mathrm{diag}\, v} U^T (\tilde{x} \otimes M_{shift}^T) g \\
+&= e^{\mathrm{diag}\, v} U^T \mathrm{vec}(M_{shift}^T g \tilde{x}^T)
 \end{align*}
 Since $U$ is the unitary map that maps a vector onto (the vectorisation of) the corresponding diagonal matrix, $U^T$ maps the vectorisation of a matrix onto its diagonal.
-If $\mathrm{Diag}: \mathbb{R}^{p \times p} \rightarrow \mathbb{p}$ maps a matrix to its diagonal, we have
-\begin{align*}
+Since the diagonal of the rank-1 matrix $M_{shift}^T g \tilde{x}^T$ is $(M_{shift}^T g) \odot \tilde{x}$, the above simplifies to
+$$
 \mathrm{grad}_v\, \left(\frac{1}{2} \norm{F(v)}^2 \right)
-&=
-e^{\mathrm{diag}\, v} \mathrm{Diag}(M_{shift}^T H \log h.(\hat{y}) \tilde{x}^T) \\
-&=
-e^{\mathrm{diag}\, v} \left((M_{shift}^T H \log h.(\hat{y})) \odot \tilde{x}\right) \\
-&=
-e^{v} \odot \left(M_{shift}^T H \log h.(\hat{y})\right) \odot \tilde{x} \\
-&=
-e^v \odot (M_{shift}^T g(\hat{y})) \odot \tilde{x}
-\end{align*}
-where $g$ is the elementwise application of $x \mapsto \log h(x) \frac{\mathrm{d}}{\mathrm{d}x}(\log h(x))$. Thus, 
-$$
-g(\hat{y}_i) = \frac{\log h(\hat{y}_i)}{h(\hat{y}_i)} \frac{\mathrm{d}}{\mathrm{d}\hat{y}_i} h(\hat{y}_i)
-$$
+=
+e^{\mathrm{diag}\, v} \left( (M_{shift}^T g) \odot \tilde{x} \right) = e^{v} \odot (M_{shift}^T g) \odot \tilde{x}
+.$$
 
-The foregoing gives the derivative of the first term in $f(W)$. The gradient of the second term is simply $\lambda v$.
+The foregoing gives the derivative of the first term in the definition of $f(v)$. The gradient of the second term is simply $\lambda v$.
+\end{proof}
 
-\end{proof}
+With this information, we can optimise \cref{eq: cost function parameterised} using any gradient-based optimisation method.
+We decided on limited-memory BFGS \cite{Nocedal2006}, which is implimented in SciPy \cite{virtanen2020scipy}.
 

### 2022-09-22 notes/notes.tex
append ideas to further work

@@
 
-To correct for non-detections, we apply the transformation $h$ from \cite{deMeutter2022uncertainty}. This is given by the continuously differentiable function
+Sometimes, the concentration at a measurement station is below the minimum detectable concentration (MDC). The corresponding measurements are $y_i = 0$.
+To account for non-detections, we apply the transformation $h$ from \cite{deMeutter2022uncertainty}. This is given by the continuously differentiable function
 $$
@@
 \begin{itemize}
+    \item Standard software libraries support optimisation with constraints. This is an alternative to regularisation and may be more interpretable.
+    Suppose that the number of components in $\tilde{x}$ is $m T$ (where $m$ is the number of sources and $T$ is the number of days).
+    If we want to bound the average of $(\log w_{ij})^2$ by some value $\delta^2$, then we have the constrained optimisation problem
+    \begin{align*}
+        \min\quad & \frac{1}{2} \norm{\log h.(\hat{y})}^2 \\
+        \text{s.t.}\quad & \frac{1}{mT} \norm{\log W}^2 \le \delta^2
+    \end{align*} 
+    which can be solved using similar ideas as those developed above.
+The prototypical constrained optimisation method (such as an active set method) would iterate towards a critical point of the Lagrangian
+    $$
+    \mathcal{L}(W, \lambda) = \frac{1}{2} \norm{\log h.(\hat{y})}^2 + \lambda\left(\frac{1}{mT} \norm{\log W}^2 - \delta^2\right)
+    $$
+    in which case the computational cost would be comparable to optimising \cref{eq: cost function regularised}.
+A more detailed explanation of constrained optimisation and the Lagrangian is found in \cite{Nocedal2006}.
+    \item Currently, there is no constraint saying that, for every emission source $j$, the geometric mean of the emission should be the reported average $x_j$.
+This can be enforced with a constraint like $\prod_{t} w_{jt} = 1$ for all sources $j=1,\dots,m$.
+With the parametrisation $w = e^v$, this is equivalent to the linear constraint $\sum_{t} v_{jt} = 0$.
+Since this is a linear constraint, it can be handled relatively easily.
     \item For certain emission sources in certain periods, there is an expected correlation between the emission at time $t$ and at time $t + 1, t + 2, \dots$.
@@
 
-    \item Standard software libraries support optimisation with constraints. This is an alternative to regularisation and may be more interpretable.
-    Suppose that the number of components in $\tilde{x}$ is $m T$ (where $m$ is the number of sources and $T$ is the number of days).
-    If we want to bound the average of $(\log w_{ij})^2$ by some value $\delta^2$, then we have the constrained optimisation problem
-    \begin{align*}
-        \min\quad & \frac{1}{2} \norm{\log h.(\hat{y})}^2 \\
-        \text{s.t.}\quad & \frac{1}{mT} \norm{\log W}^2 \le \delta^2
-    \end{align*} 
-    which can be solved using similar ideas as those developed above.
-The prototypical constrained optimisation method (such as an active set method) would iterate towards a critical point of the Lagrangian
-    $$
-    \mathcal{L}(W, \lambda) = \frac{1}{2} \norm{\log h.(\hat{y})}^2 + \lambda\left(\frac{1}{mT} \norm{\log W}^2 - \delta^2\right)
-    $$
-    in which case the computational cost would be comparable to optimising \cref{eq: cost function regularised}.
-A more detailed explanation of constrained optimisation and the Lagrangian is found in \cite{Nocedal2006}.
 \end{itemize}

### 2022-09-22 code/plot_W_timeseries.py
Plotted time series of the weights

@@ -1,0 +1,31 @@
+import numpy as np
+import constants as cst
+import matplotlib.pyplot as plt
+import constants as cst
+import pandas as pd
+
+lambdas = ["0.1", "1.0", "10.0"]
+plotsource = 200 - 1 #index of source to plot
+startindex = 380 * plotsource
+endindexp1 = 380 * (plotsource + 1)#last index + 1
+
+#copied from plot_output.py
+day0 = '20140101'
+daymin = pd.to_datetime(day0) + pd.DateOffset(-15)
+daymax = pd.to_datetime(day0) + pd.DateOffset(364)
+dates = pd.date_range(daymin, daymax)
+
+for l in lambdas:
+    string = "W_optimized_lambda_"+l+".csv"
+    #load W
+    W_opt = np.loadtxt(string)
+    W_source = W_opt[startindex:endindexp1]
+    #and reverse the indices, as the corresponding dates are ordered in reverse
+    W_source = np.flip(W_source)
+    plt.figure()
+    plt.title("Optimized weights lambda="+l)
+    plt.plot(dates, np.log10(W_source))
+    plt.xlabel("Date")
+    plt.ylabel("log10(W)")
+
+plt.show()

### 2022-09-22 notes/notes.tex
turn notes into paper

@@
 
-\title{Notes}
-\author{SCK team}
+\title{Screening of airborne radioxenon measurements for Nuclear-Test-Ban Treaty verification}
+\author{Jerome Alhage, Jelle Bosmans, Thomas Ceulemans, Nick Dewaele}
 \date{}
@@
 \newtheorem{proposition}{Proposition}
+\newtheorem{remark}[proposition]{Remark}
 
@@
 
-\section{Notation}
+    \begin{abstract}
+        We extend a simulation model for radioxenon emissions by inferring daily emissions of sources in the Northern Hemisphere.
+This is done by numerical optimisation of the relative prediction error over the daily emissions.
+    \end{abstract}
+
+\section{Introduction}
+In this paper, we adapt a simulation model of radioxenon emissions used at SCK-CEN for the verification of the Nuclear-Test-Ban Treaty.
+This model operates on a map of $m = 200$ civilian sources of radioxenon with their average daily emissions $x_1,\dots,x_m$.
+A list of $n = 4636$ measurements $[y_1,\dots,y_n]$ at stations spread across the Northern Hemisphere is given.
+The relationship between the emissions and the measured concentrations is given by a linear model
+\begin{equation}
+    \label{eq: initial model}
+y_i \approx
+\hat{y}_i = \sum_{j=1}^m \sum_{\Delta = 1}^{15}  M_{i,(j,\Delta)} \frac{x_{j}}{s_j}
+\end{equation}
+where the sensitivities $M_{i,(j, \Delta)}$ are obtained from backward simulation of the so-called \emph{xenon weather}.
+The scaling factors $s_j$ arise from the simulation on a mesh of the Northern Hemisphere.
+The \emph{sensitivity matrices} $M_{i,(j,\Delta)}$ can be interpreted as the contribution of the emission from source $j$ from $\Delta$ days before the measurement to the $i$'th concentration measurement in the dataset.
+The contributions of more than $15$ days before the measurement are neglected.
+
+\subsection{Outline}
+The above assumes that the xenon emission is constant for every source.
+We adapt this model in \cref{sec: model} so that daily emissions are taken into account.
+In \cref{sec: nondetections}, we discuss the use of a transformation from \cite{deMeutter2022uncertainty} to model the accuracy in a non-detection event.
+In \cref{sec: optimisation}, we explain a method to fit the extended model to the data.
+Given the time constraints of this project, the optimality criterion was deliberately kept simple.
+Numerical data in \cref{sec: numerical results} show that the average prediction error can be increased.
+We conclude in \cref{sec: further work} with suggestions make the model more realistic and improve computational aspects.
+
+\subsection{Notation}
 \begin{itemize}
-    \item Average emission per source: $x_1,\dots,x_m$ where $m = 200$.
-    \item Measured concentrations $y_1,\dots,y_n$ where $n = 4636$. These are consecutive measurements for several measurement stations. 
-    \item $M_{i,(j,\Delta)}$ is a block in the sensitivity matrix $M$ that represents the contribution of $x_j$ from $\Delta$ days ago to the observed concentration $y_i$.
-    \item Scaling factors $s_1,\dots,s_m$ (in \texttt{scalings.csv}). These are put in a diagonal matrix $S$.
     \item $\vb{1}_q = [1\,\, \dots\,\, 1]^T \in \mathbb{R}^q$ is the constant vector consisting of only ones.
@@
 \section{Model}
-\subsection{Initial model}
-With $n=1,\dots,4636$ and $m = 1,\dots,200$, the initial model is
-$$
-y_i \approx
-\hat{y}_i = \sum_{j=1}^m \sum_{\Delta = 1}^{15}  M_{i,(j,\Delta)} \frac{x_{j}}{s_j}
-.$$
-That is, for each emission source $j$, we simulate the contribution of 15 days, and add this together for all $j$. A matrix formulation of this reads as follows:
+\label{sec: model}
+
+\subsection{Reformulation of the baseline model}
+
+\Cref{eq: initial model} can be formulated as follows:
 \begin{align}
@@
 \left( 
-     \begin{bmatrix}
-        s_1 & & \\
-         & \ddots & \\
-        & & s_n
-     \end{bmatrix}^{-1}
+     S^{-1}
      \begin{bmatrix}
@@
 \nonumber
-.\end{align}
+\end{align}
+where $S = \mathrm{diag}([s_1,\dots,s_m])$.
+In this section, we present a different interpretation.
+Instead of $S^{-1} x \otimes \vb{1}_{15}$, we make a vector $\tilde{x} := S^{-1} x \otimes \vb{1}_{T}$ where $T = 365 + 15 = 380$.
+This is a vector of length $mT$ that gives the emission from the most recent day to $T$ days in the past for each source.
+This is assumed to be constant for all days.
+Therefore, $\tilde{x}$ consists of $m = 200$ blocks, each of which is a constant vector of length $T$.
 
-Now we present a different interpretation.
-Instead of $S^{-1} x \otimes \vb{1}_{15}$, we make a vector $\tilde{x} := S^{-1} x \otimes \vb{1}_{365}$.
-This is a vector of length $365 \times 200$ that gives the emission from day 1 to day 365 for each source.
-This is assumed to be constant for all days.
-Therefore, $\tilde{x}$ consists of $m = 200$ blocks, each of which is a constant vector of length $365$.
-
-We can write a system of equations that is logically the same as \cref{eq: initial model concise} but uses $\tilde{x}$.
+We can write a system of equations that is logically the same as \cref{eq: initial model concise} but uses $\tilde{x}$ instead of $S^{-1} x \otimes \vb{1}_{15}$.
 This gives a larger linear system.
 Because the emission from most dates has a negligible contribution to the observed concentrations, many coefficients in the larger matrix will be zero.
-For every measurement $i$, we have a linear relationship that looks something like this:
+For every measurement $i$, we have the linear relationship
 $$
@@
     \end{bmatrix}
-    (S^{-1} x \otimes \vb{1}_{365})
-.
-$$
+    \tilde{x}
+.$$
 The presence of zeros in the matrix indicates that emissions from the distant past do not contribute to the observed concentrations.
 
-Now assume that in each station, there is one observation per day, named $y_{kt}$, where $k = 1,\dots,K$ where $K$ is the number of stations and $t = 1,\dots,365$.
+Assume for the sake of simplicity that in each station, there is one observation per day, indexed as $y_{k,t}$, where $k = 1,\dots,K$ where $K$ is the number of stations and $t = 1,\dots,365$.
 The index $t=1$ corresponds to the most recent measurement.
@@
 \tilde{x}
-\end{equation}
-where 
-$$
-\tilde{x} := S^{-1} x \otimes \vb{1}_{365}
-=
-\begin{bmatrix}
-    s_1^{-1} x_1 \\
-    \vdots \\
-    s_1^{-1} x_1 \\ 
-    s_2^{-1} x_2 \\
-    \vdots \\ 
-    s_m^{-1} x_m
-\end{bmatrix}
-.$$
+.\end{equation}
 
-We could reorganise the vector $\tilde{x}$ to the transposed expression $\vb{1}_{365} \otimes (S^{-1} x)$, which is a vector containing all emissions on day 1, followed by all emissions on day 2, etc.
+\begin{remark}
+    As an alternative to using $\tilde{x} = S^{-1} x \otimes \vb{1}_T$, one could use the permuted expression $\vb{1}_{T} \otimes (S^{-1} x)$.
+This is a vector containing all emissions on day 1, followed by all emissions on day 2, etc.
 A similar permutation of $y$ could be chosen.
-This has the effect of permuting the rows and columns of $M_{shift}$.
-A permutation of the variables gives a model that is essentially the same as \cref{eq: initial model 365 days} but may have advantages when it comes to implementation or computational efficiency.
+These permutations can be compensated by the inverse permutation on the columns and rows of $M_{shift}$, respectively.
+The resulting model is essentially identical to \cref{eq: initial model 365 days} but may have advantages when it comes to implementation in software.
 We will not consider this.
+\end{remark}
 
 \subsection{Model with time-dependent correction to the emissions}
-If we weigh each day and each emission source by some factor, we are applying a transformation
+Daily emissions can be modelled by multiplying the average emission of source $j$ by some scaling factor $w_{j,t}$. This corresponds to the 
+transformation
 $$
@@
     \vdots \\
-    w_{1,365} s_1^{-1} x_1 \\ 
+    w_{1,T} s_1^{-1} x_1 \\ 
     w_{2,1} s_2^{-1} x_2 \\
     \vdots \\ 
-    w_{m, 365} s_m^{-1} x_m
+    w_{m, T} s_m^{-1} x_m
 \end{bmatrix}
 $$
-which can be written concisely as 
-$$
-W (S^{-1} x \otimes \vb{1}_{365})
-\quad\text{where}\quad 
-W = \begin{bmatrix}
-    w_{1,1} & 0 & \cdots & 0 \\
-    0 & w_{1,2} & \cdots & 0 \\
-    \vdots & \vdots & \ddots & 0 \\
-    0 & 0 & \cdots & w_{m, 365}
-\end{bmatrix}
-.$$
+which can be written concisely as $W\tilde{x}$ where $W = \mathrm{diag}([w_{1,1}, w_{1,2},\dots,w_{m,T}])$.
 Using this weighted vector of emissions instead of $\tilde{x}$ in \cref{eq: initial model 365 days} gives
@@
     \label{eq: time-evolving model concise}
-\hat{y} = M_{shift} W (S^{-1} x \otimes \vb{1}_{365})
+\hat{y} = M_{shift} W \tilde{x}
 .\end{equation}
-In this expression, the unknown is $W$ and should be fit to the data.
+In the next section, we present a method to fit the unknown factors $W$ to the data.
 
 \section{Correction for non-detections}
+\label{sec: nondetections}
 
@@
 \section{Fitting the model}
-In the above, we argued that, the prediction is accurate if $h(\hat{y})$ is close to one, where $\hat{y}$ is the predicted concentration.
+\label{sec: optimisation}
+
+In the above, we argued that the prediction is accurate if $h(\hat{y})$ is close to one, where $\hat{y}$ is the predicted concentration.
 This is equivalent to $\log^2(h(\hat{y}))$ being small.
@@
 .$$
-The average prediction error over all predictions is $\sum_{i=1}^n \log^2(h(\hat{y}_i)) = \norm{\log h(\hat{y})}^2$. This will be the first criterion for optimality of the model.
+The average prediction error over all predictions is $\sum_{i=1}^n \log^2(h(\hat{y}_i)) = \norm{\log h.(\hat{y})}^2$, where $\log h.$ is the elementwise application of $x \mapsto \log h(x)$.
+This will be the first criterion for optimality of the model.
 
@@
 
+\section{Numerical results}
+\label{sec: numerical results}
+
 \section{Further work}
+\label{sec: further work}
 \begin{itemize}

### 2022-09-23 paper/paper.tex
Added preliminary results section to paper.

--- 2022-09-22 notes/notes.tex
@@
 \usepackage{amsmath, amssymb, amsthm, dsfont, physics, hyperref, cleveref}
+\usepackage{graphicx}
 \usepackage[margin=1.5cm]{geometry}
@@
 
+Before applying any optimization procedure, we must establish a baseline.
+This is done by filling in the daily source emissions $x$ in \cref{eq: initial model} with their yearly averages.
+The simulated observations in \cref{fig: y-histogram baseline} seem to be biased towards lower values than expected given the actual measurements.\\
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.5\linewidth]{../plots/original_simulated_observation.png}
+	\caption{Distribution of the simulated observation using the yearly averaged emissions at all times.}
+	\label{fig: y-histogram baseline}
+\end{figure}
+
+In order to quantify the performance of the simulations, we define some metrics.
+First of all, the fraction of the simulated observations which lie within a factor $2$ of the actual observations.
+In the ideal case, all simulated observations $\hat{y}$ would be exactly the same as the observed ones $y$.
+This metric allows for some small scaling factor as error.
+\begin{equation}
+	\text{FAC2}:=P(0.5\leq\hat{y}/y\leq 2)
+\end{equation}
+
+The previous metric only takes into account the simulated observations within a small area.
+Therefore it cannot tell us whether some general bias is present.
+To investigate this, we use the geometric mean bias.
+The more biased the ratio is, the further this value will lie from $1$.
+This is because it defines some mean ratio of actual observations and simulated ones.
+%Nitpick Thomas: err, I did not actually check the formulae for the metrics closely during the week.
+It seems slightly weird taking the exponential, as we mostly care about the scalings (in log space).
+It might complicate the explanation: "needs to be close to 1 (in some logarithmic way)".
+Also, why do we have this minus sign in front?
+\begin{equation}
+	\text{GM}:=e^{-E(ln(\hat{y}/y))}
+\end{equation}
+
+\begin{figure}
+	\centering
+	\begin{minipage}{.48\textwidth}
+		\centering
+		\includegraphics[width=0.9\linewidth]{../plots/hfun_weights_lambda10.png}
+		\caption{Histogram of the optimized weights $W$, using a large regularization parameter $\lambda=10$.}
+		\label{fig: optimized weights lambda=10}
+	\end{minipage}\hfill
+	\begin{minipage}{.48\textwidth}
+		\centering
+		\includegraphics[width=0.9\linewidth]{../plots/hfun_simulated_observations_lambda10.png}
+		\caption{Histogram of the ratio of simulated observations $\hat{y}/y$, using a large regularization parameter $\lambda=10$.}
+		\label{fig: simulated observations lambda=10}
+	\end{minipage}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\begin{minipage}{.48\textwidth}
+		\centering
+		\includegraphics[width=0.9\linewidth]{../plots/hfun_weights_lambda1.png}
+		\caption{Histogram of the optimized weights $W$, using a medium regularization parameter $\lambda=1$.}
+		\label{fig: optimized weights lambda=1}
+	\end{minipage}\hfill
+	\begin{minipage}{.48\textwidth}
+		\centering
+		\includegraphics[width=0.9\linewidth]{../plots/hfun_simulated_observations_lambda1.png}
+		\caption{Histogram of the ratio of simulated observations $\hat{y}/y$, using a medium regularization parameter $\lambda=1$.}
+		\label{fig: simulated observations lambda=1}
+	\end{minipage}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\begin{minipage}{.48\textwidth}
+		\centering
+		\includegraphics[width=0.9\linewidth]{../plots/hfun_weights_lambda01.png}
+		\caption{Histogram of the optimized weights $W$, using a small regularization parameter $\lambda=0.1$.}
+		\label{fig: optimized weights lambda=0.1}
+	\end{minipage}\hfill
+	\begin{minipage}{.48\textwidth}
+		\centering
+		\includegraphics[width=0.9\linewidth]{../plots/hfun_simulated_observations_lambda01.png}
+		\caption{Histogram of the ratio of simulated observations $\hat{y}/y$, using a small regularization parameter $\lambda=0.1$.}
+		\label{fig: simulated observations lambda=0.1}
+	\end{minipage}
+\end{figure}
+
+Starting with the results when using a high regularization parameter $\lambda=10$, we observe that the weights are nearly $1$ everywhere (see \cref{fig: optimized weights lambda=10}).
+This makes sense, as the cost function in this case heavily penalizes the weights for deviating from $1$.
+As a result, the histogram of simulated results (see \cref{fig: simulated observations lambda=10}) is nearly identical to the baseline.
+Lowering the regularization parameter $\lambda=1$, we find that the weights have more freedom to deviate from $1$ (see \cref{fig: optimized weights lambda=1}) and the bias on the ratio $\hat{y}/y$ decreases (see \cref{fig: simulated observations lambda=1}).
+This is a direct result of lowering the influence of the regularization term.
+Finally, lowering the regularization parameter $\lambda=0.1$ even further, the weights get even more freedom (see \cref{fig: optimized weights lambda=0.1}), while the ratio $\hat{y}/y$ finally seems to be centered around $1$.
+In \cref{table: evaluated metrics}, a summary of the metrics is found, evaluated at the different values for the regularization parameter.
+Summarized, we see the most major change in the geometric mean bias GM as the regularization parameter decreases.
+This might lead us to believe that the best result is obtained by choosing $\lambda$ as small as possible.
+However, a small value for $\lambda$ results in larger deviations from the mean emission; these can become greater than can what be expected in real life TODO REF PAPER typical variation.
+Thus a fine balance between the prediction error and the regularization condition is still needed.\\
+
+\begin{table}
+	\centering
+	\caption{Metrics applied to the simulated observations in case of different regularization parameters $\lambda$.}\label{table: evaluated metrics}
+\begin{tabular}{|l||c|c|}
+	\hline
+	&FAC2&GM\\\hline
+	baseline&$31.7\%$&3.050\\\hline
+	$\lambda=10$&$32.2\%$&2.984\\\hline	
+	$\lambda=1$&$40.5\%$&1.776\\\hline	
+	$\lambda=0.1$&$40.8\%$&1.216\\\hline
+\end{tabular}
+\end{table}
+
 \section{Further work}

### 2022-10-14 paper/paper.tex
Added non-technical intro, references, fixed some typos/notation

@@
 \section{Introduction}
-In this paper, we adapt a simulation model of radioxenon emissions used at SCK-CEN for the verification of the Nuclear-Test-Ban Treaty.
+The Comprehensive Nuclear Test Ban Treaty \cite{CTBT} bans all atmospheric, underwater and underground nuclear explosions when it enters into force.
+Compliance with the Treaty will be verified by the International Monitoring System to which SCK-CEN contributes as a National Data Centre.
+As part of the monitoring system, more than 300 measuring stations have been set up worldwide including seismic, hydro-acoustic, infrasound and radionuclide detectors.
+In this paper, we focus on the radionuclide detectors and in particular on measurements of radioactive xenon or \emph{radioxenon} concentrations.
+Monitoring of radioxenon concentrations is particularly useful for the detection of underground nuclear explosions, since radioxenon readily escapes the explosion zone owning to its properties as a noble gas.
+However, civilian sources such as medical isotope production facilities and nuclear power plants also emit radioactive xenon and this \emph{civilian background} complicates the detection of nuclear explosions.
+
+One approach under study at SCK-CEN explicitly models the \emph{xenon weather} using emission data from civilian sources and atmospheric transport modelling.
+If this approach can be used to accurately predict the civilian background, emissions caused by nuclear explosions can be separated from civilian sources.
+In this paper, we adapt the simulation model of radioxenon emissions and transport developped at SCK-CEN.
 This model operates on a map of $m = 200$ civilian sources of radioxenon with their average daily emissions $x_1,\dots,x_m$.
 A list of $n = 4636$ measurements $[y_1,\dots,y_n]$ at stations spread across the Northern Hemisphere is given.
-The relationship between the emissions and the measured concentrations is given by a linear model
+The relationship between the emissions and the measured concentrations is given by a linear model,
 \begin{equation}
@@
 \end{equation}
-where the sensitivities $M_{i,(j, \Delta)}$ are obtained from backward simulation of the so-called \emph{xenon weather}.
+where the sensitivities $M_{i,(j, \Delta)}$ are obtained from backward simulation of the atmospheric transport.
 The scaling factors $s_j$ arise from the simulation on a mesh of the Northern Hemisphere.
+In particular, they compensate for the fact that not all cells in the mesh have the same volume.
 The \emph{sensitivity matrices} $M_{i,(j,\Delta)}$ can be interpreted as the contribution of the emission from source $j$ from $\Delta$ days before the measurement to the $i$'th concentration measurement in the dataset.
 The contributions of more than $15$ days before the measurement are neglected.
+A certain source will only contribute to the measurements that are relatively close to it, depending on the prevailing wind directions that facilitate atmospheric transport.
+Hence, many sensitivity matrices will contain zeros when there is no transport from the source to the measurement location.
 
@@
 In \cref{sec: nondetections}, we discuss the use of a transformation from \cite{deMeutter2022uncertainty} to model the accuracy in a non-detection event.
-In \cref{sec: optimisation}, we explain a method to fit the extended model to the data.
+In \cref{sec: optimisation}, we explain a method to fit the extended model to the measured concentration data.
 Given the time constraints of this project, the optimality criterion was deliberately kept simple.
-Numerical data in \cref{sec: numerical results} show that the average prediction error can be increased.
-We conclude in \cref{sec: further work} with suggestions make the model more realistic and improve computational aspects.
+Numerical data in \cref{sec: numerical results} show that the average prediction error can be decreased.
+We conclude in \cref{sec: further work} with suggestions to make the model more realistic and improve computational aspects.
 
@@
     \item $\odot$ is the Hadamard (i.e.\, elementwise) product, given by $(AB)_{ij} = A_{ij} B_{ij}$.
-    \item $\mathrm{vec}: \mathbb{R}^{p \times q} \rightarrow \mathbb{R}^{pq}$ is the vectorisation operator. It can be understood as stacking the columns of a matrix in one long vector. 
+    \item $\mathrm{vec}(\bullet): \mathbb{R}^{p \times q} \rightarrow \mathbb{R}^{pq}$ is the vectorisation operator. It can be understood as stacking the columns of a matrix in one long vector. 
     \item $\mathrm{diag}$ is the canonical identification of $p$-vectors and $p \times p$ diagonal matrices, i.e.,
@@
         A standard rule of thumb in numerical linear algebra is that the Kronecker product serves a mostly theoretical purpose and is rarely calculated explicitly.
-        This is because the Kronecker product satisfies $\mathrm{vec} (AXB) = (B^T \otimes A) \mathrm{vec} X$ for all matrices $A,X,B$ \cite[eq. 1.3.6]{Golub2013}.
+        This is because the Kronecker product satisfies $\mathrm{vec} \left(AXB\right) = \left(B^T \otimes A\right) \mathrm{vec}\left(X\right)$ for all matrices $A,X,B$ \cite[eq. 1.3.6]{Golub2013}.
     \item For a vector-valued function $F(x)$, the Jacobian matrix is denoted as $\frac{\partial F}{\partial x}$.
@@
 The presence of zeros in the matrix indicates that emissions from the distant past do not contribute to the observed concentrations.
+Furthermore, recall many of the sensitivity matrices themselves also contain zeros.
+Hence the larger matrix has a sparse nature.
 
@@
 \begin{proof}
-Let $W = e^{\mathrm{diag}\, v}$ and let $U$ be the matrix corresponding to the linear map $\mathrm{vec} \circ \mathrm{diag}$. Then $\mathrm{vec}\, W(v) = U e^v$.
+Let $W = e^{\mathrm{diag}\, v}$ and let $U$ be the matrix corresponding to the linear map $\mathrm{vec} \circ \mathrm{diag}$. Then $\mathrm{vec}\left(W(v)\right) = U e^v$.
 Define $$
@@
 =
-(\tilde{x}^T \otimes M_{shift}) \mathrm{vec} W(v)
+(\tilde{x}^T \otimes M_{shift}) \mathrm{vec} \left(W(v)\right)
 =
@@
 &= e^{\mathrm{diag}\, v} U^T (\tilde{x} \otimes M_{shift}^T) g \\
-&= e^{\mathrm{diag}\, v} U^T \mathrm{vec}(M_{shift}^T g \tilde{x}^T)
+&= e^{\mathrm{diag}\, v} U^T \mathrm{vec}\left(M_{shift}^T g \tilde{x}^T\right)
 \end{align*}
@@
 With this information, we can optimise \cref{eq: cost function parameterised} using any gradient-based optimisation method.
-We decided on limited-memory BFGS \cite{Nocedal2006}, which is implimented in SciPy \cite{virtanen2020scipy}.
+We decided on limited-memory BFGS \cite{Nocedal2006}, which is implemented in SciPy \cite{virtanen2020scipy}.
 
@@
 This might lead us to believe that the best result is obtained by choosing $\lambda$ as small as possible.
-However, a small value for $\lambda$ results in larger deviations from the mean emission; these can become greater than can what be expected in real life TODO REF PAPER typical variation.
+However, a small value for $\lambda$ results in larger deviations from the mean emission; these can become greater than can what be expected in real life typical variation \cite{Ringbom2021,DeMeutter2016}.
 Thus a fine balance between the prediction error and the regularization condition is still needed.\\

### 2022-10-21 paper/paper.tex
Added time series plot and accompanying text.

@@
 
+To grant some intuition about the time dependence of the variability of the optimized emissions, the time series of the weights are plotted for different values of the regularization parameter (see \cref{fig: optimized weights timeseries lambda=10}, \cref{fig: optimized weights timeseries lambda=1}, \cref{fig: optimized weights timeseries lambda=0.1}).
+At first glance, all time series look erratic.
+Looking at the measured emission data in \cite{Ringbom2021}, the data is seemingly also erratic.
+This can support the claim that our optimized emissions could be realistic, however proper time series analysis is still needed to extract firm conclusions.
+At the same time, we see that all time series for the optimized weights look similar (ignoring the scale differences).
+This can be expected, as all optimized solutions try to fit the same data but are hindered in different amounts by their respective regularization parameter.
+
+\begin{figure}
+	\centering
+	\begin{minipage}{.31\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{../plots/Optimized_weights_timeseries_lambda10.png}
+		\caption{Time series of the optimized weights $W$, using a large regularization parameter $\lambda=10$.}
+		\label{fig: optimized weights timeseries lambda=10}
+	\end{minipage}\hfill
+	\begin{minipage}{.31\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{../plots/Optimized_weights_timeseries_lambda1.png}
+		\caption{Time series of the optimized weights $W$, using a medium regularization parameter $\lambda=1$.}
+		\label{fig: optimized weights timeseries lambda=1}
+	\end{minipage}\hfill
+	\begin{minipage}{.31\textwidth}
+	\centering
+	\includegraphics[width=\linewidth]{../plots/Optimized_weights_timeseries_lambda01.png}
+	\caption{Time series of the optimized weights $W$, using a small regularization parameter $\lambda=0.1$.}
+	\label{fig: optimized weights timeseries lambda=0.1}
+	\end{minipage}
+\end{figure}
+
 \begin{table}

### 2022-10-24 paper/paper.tex
add anomaly detection section

@@
 
+\section{Anomaly detection}
+\begin{itemize}
+
+The main challenges of applying traditional anomaly detection methods, on time-series data of an observation station, is accommodating for weather.
+Winds will change which observation station an emission will end up in, therefore treating stations individually to try to learn their normal behavior would not lead to fruitful results.
+At the same time, it is also difficult to train more complex model that handles data collectively from all stations, since there is not enough historic data, and few observation stations are collecting data throughout the entire year.
+
+This is why working with the emission sources is more appealing.
+For the observation stations that are mostly operational during the year, correcting the emission data in the previous chapters allows us to obtain a better ground truth for observed emissions.
+The individual time-series data should show a few characteristics that arise from the nature of these sources:
+    \item For power plants, a long-term baseline variation corresponding to yearly electricity demand, which should vary consistently given that our readings are all in the northern hemisphere.
+    \item For medical isotope production facilities, a short term periodic trend corresponding to the working days of the week.
+
+After adjusting for the yearly tend, the eventual solution should include two components.
+The first to segment the time-series into pieces with distinct behavior or features, for example using shapelets \cite{10.1145/2623330.2623613},\cite{10.1007/978-3-030-39098-3_7} or change point detection algorithms \cite{TRUONG2020107299}.
+The second is to train a 1D anomaly detection algorithm that considers these segments as normal behavior.
+The state of the art methods use convolution neural networks \cite{KIRANYAZ2021107398} or bayesian probabilistic models \cite{https://doi.org/10.48550/arxiv.1902.08627}.
+Finally, synthetic emission data should be simulated in order to evaluate these methods, since data for nuclear tests is rare if not inexistant.
+
+\end{itemize}
+
 \printbibliography

### 2022-10-24 paper/paper.tex
Fixed compilation of anomaly detection section.

@@
 \section{Anomaly detection}
-\begin{itemize}
 
@@
 Winds will change which observation station an emission will end up in, therefore treating stations individually to try to learn their normal behavior would not lead to fruitful results.
-At the same time, it is also difficult to train more complex model that handles data collectively from all stations, since there is not enough historic data, and few observation stations are collecting data throughout the entire year.
+At the same time, it is also difficult to train more complex model that handles data collectively from all stations, since there is not enough historic data, and few observation stations are collecting data throughout the entire year.\\
 
@@
 The individual time-series data should show a few characteristics that arise from the nature of these sources:
+\begin{itemize}
     \item For power plants, a long-term baseline variation corresponding to yearly electricity demand, which should vary consistently given that our readings are all in the northern hemisphere.
     \item For medical isotope production facilities, a short term periodic trend corresponding to the working days of the week.  
-
-After adjusting for the yearly tend, the eventual solution should include two components.
+\end{itemize}  
+After adjusting for the yearly trend, the eventual solution should include two components.
 The first to segment the time-series into pieces with distinct behavior or features, for example using shapelets \cite{10.1145/2623330.2623613},\cite{10.1007/978-3-030-39098-3_7} or change point detection algorithms \cite{TRUONG2020107299}.
@@
 
-\end{itemize}
-
 \printbibliography

### 2022-10-24 paper/paper.tex
fix grammar anomaly detection

--- 2022-10-21 paper/paper.tex
@@
 
+\section{Anomaly detection}
+
+One of the main challenges of applying traditional anomaly detection methods on the time-series data of an observation station, is accommodating for weather.
+Winds will change the course of an emission, causing different observation stations to pick up their signature.
+Therefore treating stations individually to try to learn their normal behavior would not lead to fruitful results.
+At the same time, it is also difficult to train more complex models that handle data collectively from all stations, since there is not enough historic data, and only a few observation stations are actively collecting data throughout the entire year.\\
+
+For these reasons, working with the emission sources directly is more appealing.
+Correcting the emission data in the previous chapters allows us to obtain a better ground truth for the variation of the emissions over the course of the year.
+The resulting individual time-series data should show a few characteristics that arise from the nature of these sources:
+\begin{itemize}
+    \item For power plants, a long-term baseline variation corresponding to yearly electricity demand, which should vary consistently given that our readings are all in the northern hemisphere.
+    \item For medical isotope production facilities, a short term periodic trend corresponding to the working days of the week.  
+\end{itemize}
+
+After adjusting for the yearly trend, the eventual solution should include two components.
+The first which will segment the time-series into pieces with distinct behavior or features, for example using shapelets \cite{10.1145/2623330.2623613}\cite{10.1007/978-3-030-39098-3_7} or change point detection algorithms \cite{TRUONG2020107299}.
+The second part is to train anomaly detection models or classifiers, that distinguish between segments of normal and abnormal behavior.
+The state of the art methods use 1D convolution neural networks \cite{KIRANYAZ2021107398} or bayesian probabilistic models \cite{https://doi.org/10.48550/arxiv.1902.08627}.
+Finally, synthetic emission data should be simulated in order to evaluate these methods, since data for nuclear tests is rare if not inexistent.
+
 \printbibliography

### 2022-10-26 paper/paper.tex
Moved anomaly detection chapter before conclusions and added it to the outline Minor corrections and clarifications

@@
 \subsection{Outline}
-The above assumes that the xenon emission is constant for every source.
+The model above assumes that the xenon emission is constant for every source.
 We adapt this model in \cref{sec: model} so that daily emissions are taken into account.
@@
 Numerical data in \cref{sec: numerical results} show that the average prediction error can be decreased.
-We conclude in \cref{sec: further work} with suggestions to make the model more realistic and improve computational aspects.
+In \cref{sec:anomaly detection} we discuss how anomaly detection could be applied on the resulting time-corrected emissions.
+We conclude in \cref{sec: further work} with suggestions to make the model more realistic, and improve computational aspects.
 
@@
 This is a vector of length $mT$ that gives the emission from the most recent day to $T$ days in the past for each source.
-This is assumed to be constant for all days.
+This is assumed to be constant for all days (for now).
 Therefore, $\tilde{x}$ consists of $m = 200$ blocks, each of which is a constant vector of length $T$.
@@
 These permutations can be compensated by the inverse permutation on the columns and rows of $M_{shift}$, respectively.
-The resulting model is essentially identical to \cref{eq: initial model 365 days} but may have advantages when it comes to implementation in software.
+The resulting model is essentially identical to \cref{eq: initial model 365 days} but may have advantages when it comes to implementation in software (e.g.
+sparsity structure).
 We will not consider this.
@@
 \subsection{Model with time-dependent correction to the emissions}
-Daily emissions can be modelled by multiplying the average emission of source $j$ by some scaling factor $w_{j,t}$. This corresponds to the 
+Daily emission variations can be modelled by multiplying the average emission of source $j$ by some scaling factor $w_{j,t}$. This corresponds to the 
 transformation
@@
 .\end{equation}
-In the next section, we present a method to fit the unknown factors $W$ to the data.
+In the next section, we present a method to fit the unknown factors $w_{j,t}$ to the data.
 
@@
 
-Sometimes, the concentration at a measurement station is below the minimum detectable concentration (MDC). The corresponding measurements are $y_i = 0$.
+Sometimes, the concentration at a measurement station is below the Minimum Detectable Concentration (MDC). The corresponding measurements are $y_i = 0$.
 To account for non-detections, we apply the transformation $h$ from \cite{deMeutter2022uncertainty}. This is given by the continuously differentiable function
@@
 the factors $W$.
+This will act as a regularisation term that keeps the daily emissions from varying wildly.
 In a similar vein as the above, the size of $W$ can be measured as $\norm{\log W}^2$ where $\norm{\cdot}$ is the Frobenius norm\footnote{Since $W$ is a diagonal matrix, the matrix logarithm $\log W$ is simply the diagonal matrix with $\log w_{1,1}, \log w_{1,2}, \dots$ on the diagonal.
@@
 
-The previous metric only takes into account the simulated observations within a small area.
+The previous metric only takes into account the statistical spread of simulated observations within a small area.
 Therefore it cannot tell us whether some general bias is present.
@@
 
+\section{Anomaly detection}
+\label{sec:anomaly detection}
+One of the main challenges of applying traditional anomaly detection methods on the time-series data of an observation station, is accommodating for weather.
+Winds will change the course of an emission, causing different observation stations to pick up their signature.
+Therefore treating stations individually to try to learn their normal behavior would not lead to fruitful results.
+At the same time, it is also difficult to train more complex models that handle data collectively from all stations, since there is not enough historic data, and only a few observation stations are actively collecting data throughout the entire year.\\
+
+For these reasons, working with the emission sources directly is more appealing.
+Correcting the emission data in the previous chapters allows us to obtain a better ground truth for the variation of the emissions over the course of the year.
+The resulting individual time-series data should show a few characteristics that arise from the nature of these sources:
+\begin{itemize}
+	\item For power plants, a long-term baseline variation corresponding to yearly electricity demand, which should vary consistently given that our readings are all in the northern hemisphere.
+	\item For medical isotope production facilities, a short term periodic trend corresponding to the working days of the week.  
+\end{itemize}
+
+After adjusting for the yearly trend, the eventual solution should include two components.
+The first which will segment the time-series into pieces with distinct behavior or features, for example using shapelets \cite{10.1145/2623330.2623613}\cite{10.1007/978-3-030-39098-3_7} or change point detection algorithms \cite{TRUONG2020107299}.
+The second part is to train anomaly detection models or classifiers, that distinguish between segments of normal and abnormal behavior.
+The state of the art methods use 1D convolution neural networks \cite{KIRANYAZ2021107398} or bayesian probabilistic models \cite{https://doi.org/10.48550/arxiv.1902.08627}.
+Finally, synthetic emission data should be simulated in order to evaluate these methods, since data for nuclear tests is rare if not inexistent.
+
 \section{Further work}
@@
 
-\section{Anomaly detection}
-
-One of the main challenges of applying traditional anomaly detection methods on the time-series data of an observation station, is accommodating for weather.
-Winds will change the course of an emission, causing different observation stations to pick up their signature.
-Therefore treating stations individually to try to learn their normal behavior would not lead to fruitful results.
-At the same time, it is also difficult to train more complex models that handle data collectively from all stations, since there is not enough historic data, and only a few observation stations are actively collecting data throughout the entire year.\\
-
-For these reasons, working with the emission sources directly is more appealing.
-Correcting the emission data in the previous chapters allows us to obtain a better ground truth for the variation of the emissions over the course of the year.
-The resulting individual time-series data should show a few characteristics that arise from the nature of these sources:
-\begin{itemize}
-    \item For power plants, a long-term baseline variation corresponding to yearly electricity demand, which should vary consistently given that our readings are all in the northern hemisphere.
-    \item For medical isotope production facilities, a short term periodic trend corresponding to the working days of the week.  
-\end{itemize}
-
-After adjusting for the yearly trend, the eventual solution should include two components.
-The first which will segment the time-series into pieces with distinct behavior or features, for example using shapelets \cite{10.1145/2623330.2623613}\cite{10.1007/978-3-030-39098-3_7} or change point detection algorithms \cite{TRUONG2020107299}.
-The second part is to train anomaly detection models or classifiers, that distinguish between segments of normal and abnormal behavior.
-The state of the art methods use 1D convolution neural networks \cite{KIRANYAZ2021107398} or bayesian probabilistic models \cite{https://doi.org/10.48550/arxiv.1902.08627}.
-Finally, synthetic emission data should be simulated in order to evaluate these methods, since data for nuclear tests is rare if not inexistent.
-
 \printbibliography
